---
title: "BC population and geography data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BC population and geography data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(dplyr)
library(islandepi)
```

`islandepi` provides a controlled access layer around the BC Data Catalogue.
Permanent record and resource identifiers are pinned in the package, downloaded
objects use consistent join keys, and no request is made when the package loads.

## Confirm the pinned sources

`islh_bc_sources()` does not make a network request. It lists the exact
catalogue records and resources the package will use.

```{r sources}
islh_bc_sources()
```

## Population denominators

The population helper supports Local Health Areas and Health Service Delivery
Areas. It reshapes one-year ages into the same age bands used by
`islh_age_group()`.

```{r download-population, eval=FALSE}
lha_population <- islh_bc_population(
  geography = "lha",
  years = 2021:2026,
  sex = c("F", "M"),
  age_breaks = "five_year"
)
```

Female and male rows are returned separately. Summing them produces a total
without double counting. Alternatively, request `sex = "T"` on its own when
the catalogue total is required.

The downloaded object has eight columns. The rows below are typed in so the
website can show the shape without a live catalogue request. The values are
BC Stats' 2025 estimates for two Island Health LHAs, both sexes, from the
Local Health Areas resource of the BC Sub-Provincial Population Estimates and
Projections record. To keep the example short, it shows only the first three
five-year age bands, ages 0 to 14. A real extract has every age band.

```{r population-shape}
population_example <- data.frame(
  geography = rep("lha", 6),
  geography_code = c("411", "411", "411", "412", "412", "412"),
  geography_name = c(
    rep("Greater Victoria", 3),
    rep("Western Communities", 3)
  ),
  year = rep(2025L, 6),
  estimate_type = rep("Estimate", 6),
  sex = rep("T", 6),
  age_group = factor(
    rep(c("0-4", "5-9", "10-14"), 2),
    levels = c("0-4", "5-9", "10-14"),
    ordered = TRUE
  ),
  population = c(7313, 9349, 10222, 5054, 6155, 6244)
)
population_example
```

Summing over age groups gives one denominator per geography, but only for the
ages present. Check which bands a sum covers before you name it. Here the sum
is the population aged 0 to 14, not the total population, and the object is
named to say so:

```{r population-total}
unique(population_example$age_group)

population_0_14 <- population_example |>
  summarise(
    population = sum(population),
    .by = c(geography, geography_code, geography_name, year, estimate_type)
  )
population_0_14
```

Use a denominator like this only with a numerator restricted to the same ages,
such as cases aged 0 to 14. An all-ages rate needs the sum over every age band
of the full extract.

## Health-geography boundaries

The geography helper supports health authorities, HSDAs, LHAs and CHSAs. It
returns an `sf` object in BC Albers by default and limits the result to Island
Health.

```{r download-boundaries, eval=FALSE}
lha_boundaries <- islh_bc_geography("lha")

# Use all provincial boundaries when needed.
bc_chsa <- islh_bc_geography("chsa", health_authority = NULL)
```

`health_authority = "Island Health"` and `"Vancouver Island"` select the same
catalogue records.

The returned object contains these analysis columns plus its `sf` geometry:

| Column | Meaning | Example |
|---|---|---|
| `geography` | Stable geography type | `"lha"` |
| `geography_code` | Zero-padded join key | `"414"` |
| `geography_name` | Display name | `"Southern Gulf Islands"` |
| `health_authority_name` | Catalogue authority name | `"Vancouver Island"` |
| `geometry` | Boundary in the requested CRS | `MULTIPOLYGON` |

The catalogue boundaries are detailed. For a report map, simplify them with
`rmapshaper::ms_simplify(keep_shapes = TRUE)` rather than the
`simplify_tolerance` argument, so neighbouring areas keep a shared border
instead of opening slivers between them. For a map that works offline,
`islandbrand::islh_example_lha()` returns a simplified copy of the 14 Island
Health LHAs with 2025 populations.

## A checked population join

Use the standardized geography type and code for a transparent join. The
population resource covers all of BC, while the boundary helper defaults to
Island Health. Extra population rows are therefore expected to be discarded.
The one-to-one relationship still rejects duplicate geography codes.

```{r checked-live-join, eval=FALSE}
lha_extract <- islh_bc_population("lha", years = 2025, sex = "T")
lha_source <- attr(lha_extract, "bcdata_source")

lha_total <- lha_extract |>
  summarise(
    population = sum(population),
    .by = c(geography, geography_code, year, estimate_type)
  )

lha_boundaries <- islh_bc_geography("lha")

lha_map_data <- left_join(
  lha_boundaries,
  lha_total,
  by = join_by(geography, geography_code),
  relationship = "one-to-one",
  na_matches = "never"
)

if (anyNA(lha_map_data$population)) {
  stop("One or more Island Health boundaries did not match population data.")
}
```

Use the boundary object on the left of the join. Population data cover all of
BC, while the default boundaries contain only Island Health. A join that
requires every population row to match will therefore fail by design.

For a table of counts rather than boundaries, `islh_join_denominator()` makes
the same checks in one step. It stops if any count has no denominator, if the
denominator table has more than one row per key, or if a code is text in one
table and a number in the other:

```{r checked-example-join}
cases_0_14 <- data.frame(
  geography = c("lha", "lha"),
  geography_code = c("411", "412"),
  cases = c(38, 21)
)

joined_example <- islh_join_denominator(
  cases_0_14,
  population_0_14,
  by = c("geography", "geography_code")
)
joined_example
```

The case counts are illustrative. When the join stops, check
code padding, geography type, year and estimate type before calculating.
Population rows with no matching count, such as the rest of BC, are reported
in a message, since a stratum with zero cases still belongs in a rate.

Read the `bcdata_source` attribute immediately after download, or use
`islh_bc_sources()`, to record catalogue provenance. Attributes may be dropped
by later data transformations, so persist source metadata explicitly when it
must travel with an analytical extract.

## Reproducible extracts

The package retrieves current catalogue content and deliberately does not ship
copied datasets. For a governed analysis, save the retrieved object in the
project with its retrieval date and source metadata. Re-run the retrieval only
when the analysis protocol permits the denominator or geography version to
change.

If a request fails on a health-authority network, verify that
`catalogue.data.gov.bc.ca` is reachable through the configured proxy. The
package does not modify `bcdata` internals or bypass its connectivity checks.

## Common questions

**Why did requesting `sex = "T"` still return F and M rows?** Update
`islandepi` and restart R so the installed namespace matches the current
package. Then inspect `unique(result$sex)`.

**Why are there more population areas than boundaries?** Population resources
cover the province. The default boundary query filters to Island Health.

**Why did a one-to-one join reject the data?** Population output has one row
per age group and sex. Aggregate to one row per geography first, as shown
above, summing over every age band you mean to include.

**How do I reproduce a published map later?** Save the downloaded objects,
their `bcdata_source` attributes and the extraction date in the analysis
project.
