---
title: "BC population and geography data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BC population and geography data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(dplyr)
library(islandepi)
```

`islandepi` provides a controlled access layer around the BC Data Catalogue.
Permanent record and resource identifiers are pinned in the package, downloaded
objects use consistent join keys, and no request is made when the package loads.

## Confirm the pinned sources

`islh_bc_sources()` does not make a network request. It lists the exact
catalogue records and resources the package will use.

```{r sources}
islh_bc_sources()
```

## Population denominators

The population helper supports Local Health Areas and Health Service Delivery
Areas. It reshapes one-year ages into the same age bands used by
`islh_age_group()`.

```{r download-population, eval=FALSE}
island_lha_codes <- c(
  "411", "412", "413", "414", "421", "422", "423", "424", "425", "426",
  "431", "432", "433", "434"
)

lha_population <- islh_bc_population(
  geography = "lha",
  years = 2021:2025,
  sex = c("F", "M"),
  age_breaks = "five_year",
  estimate_type = "Estimate",
  expected_codes = island_lha_codes
)
```

`expected_codes` comes from your own roster, not from the download. If BC
Stats renames, merges or drops a local health area, the request stops and
names the missing code, rather than letting a later join quietly lose the
area. The resource holds estimates up to the latest year and projections after
it; `estimate_type` says which you mean.

Female and male rows are returned separately. Summing them produces a total
without double counting. Alternatively, request `sex = "T"` on its own when
the catalogue total is required.

### What the estimates describe

Each estimate is the population on **1 July** of its year, so it is a
mid-year value: the usual denominator for a rate over that calendar year. For
a period that crosses two years, such as a respiratory season from September
to April, either use the year that holds most of the period or average the
two years, and say which in the release. For a period shorter than a year, say
the rate is for that period, as the [Getting started](islandepi.html) article
does for an 18-week window, or convert the population to person-time.

BC Stats labels the sex column `Gender`, with the categories `F` and `M`;
`islh_bc_population()` names it `sex`. For every area the two sum to the
published total, `T`. That is the catalogue's definition, not necessarily the
one in your case data. Before calculating rates by sex, check how the line
list records sex or gender, and how it codes people who are neither `F` nor
`M` or whose value is missing. Cases in a category the denominator does not
have cannot go into a sex-specific rate, so report how many there are.

### The shape of an extract

The downloaded object has eight columns: `geography`, `geography_code`,
`geography_name`, `year`, `estimate_type`, `sex`, `age_group` and
`population`. For examples and teaching, the package also holds a saved copy
of the 2025 estimates for Island Health's 14 local health areas,
`islh_lha_population`. It has the same columns, except that `hsda` takes the
place of `geography`, and it needs no network request. Use a fresh download
for a report.

To keep the example short, it takes two local health areas and the first
three five-year age bands, ages 0 to 14:

```{r population-shape}
population_example <- islh_lha_population |>
  filter(
    geography_name %in% c("Greater Victoria", "Western Communities"),
    age_group <= "10-14"
  )
population_example
```

Summing over sex and age group gives one denominator per area, but only for
the ages present. Check which bands a sum covers before you name it. Here the
sum is the population aged 0 to 14, not the total population, and the object
is named to say so:

```{r population-total}
population_example |>
  distinct(age_group)

population_0_14 <- population_example |>
  summarise(
    population = sum(population),
    .by = c(geography_code, geography_name, year, estimate_type)
  )
population_0_14
```

Use a denominator like this only with a numerator restricted to the same ages,
such as cases aged 0 to 14. An all-ages rate needs the sum over every age band
of the full extract.

## Health-geography boundaries

The geography helper supports health authorities, HSDAs, LHAs and CHSAs. It
returns an `sf` object in BC Albers by default and limits the result to Island
Health.

```{r download-boundaries, eval=FALSE}
lha_boundaries <- islh_bc_geography("lha")

# Use all provincial boundaries when needed.
bc_chsa <- islh_bc_geography("chsa", health_authority = NULL)
```

`health_authority = "Island Health"` and `"Vancouver Island"` select the same
catalogue records.

The returned object contains these analysis columns plus its `sf` geometry:

| Column | Meaning | Example |
|---|---|---|
| `geography` | Stable geography type | `"lha"` |
| `geography_code` | Zero-padded join key | `"414"` |
| `geography_name` | Display name | `"Southern Gulf Islands"` |
| `health_authority_name` | Catalogue authority name | `"Vancouver Island"` |
| `geometry` | Boundary in the requested CRS | `MULTIPOLYGON` |

The catalogue boundaries are detailed. For a report map, simplify them with
`rmapshaper::ms_simplify(keep_shapes = TRUE)` rather than the
`simplify_tolerance` argument, so neighbouring areas keep a shared border
instead of opening slivers between them. For a map that works offline,
`islandbrand::islh_example_lha()` returns a simplified copy of the 14 Island
Health LHAs with 2025 populations.

## A checked population join

Use the standardized geography type and code for a transparent join. The
population resource covers all of BC, while the boundary helper defaults to
Island Health. Extra population rows are therefore expected to be discarded.
The one-to-one relationship still rejects duplicate geography codes.

```{r checked-live-join, eval=FALSE}
lha_extract <- islh_bc_population("lha", years = 2025, sex = "T")
lha_source <- attr(lha_extract, "bcdata_source")

lha_total <- lha_extract |>
  summarise(
    population = sum(population),
    .by = c(geography, geography_code, year, estimate_type)
  )

lha_boundaries <- islh_bc_geography("lha")

lha_map_data <- left_join(
  lha_boundaries,
  lha_total,
  by = join_by(geography, geography_code),
  relationship = "one-to-one",
  na_matches = "never"
)

if (anyNA(lha_map_data$population)) {
  stop("One or more Island Health boundaries did not match population data.")
}
```

Use the boundary object on the left of the join. Population data cover all of
BC, while the default boundaries contain only Island Health. A join that
requires every population row to match will therefore fail by design.

For a table of counts rather than boundaries, `islh_join_denominator()` makes
the same checks in one step. It stops if any count has no denominator, if the
denominator table has more than one row per key, or if a code is text in one
table and a number in the other:

```{r checked-example-join}
cases_0_14 <- tibble(
  geography_code = c("411", "412"),
  cases = c(38, 21)
)

cases_0_14 |>
  islh_join_denominator(population_0_14, by = "geography_code") |>
  mutate(
    islh_crude_rate(cases, population, per = 100000) |>
      select(rate, lower, upper)
  )
```

The case counts are illustrative. When the join stops, check
code padding, geography type, year and estimate type before calculating.
Population rows with no matching count, such as the rest of BC, are reported
in a message, since a stratum with zero cases still belongs in a rate.

Read the `bcdata_source` attribute immediately after download, or use
`islh_bc_sources()`, to record catalogue provenance. Attributes may be dropped
by later data transformations, so persist source metadata explicitly when it
must travel with an analytical extract.

## Reproducible extracts

The download functions retrieve current catalogue content. `islh_lha_population`
is a fixed copy for examples, not a substitute. The pinned record and resource
identifiers say where the data came from, but not which version: BC Stats revises estimates, and a later
download of the same resource can differ. For a governed analysis, save the
retrieved object in the project with its retrieval date and source metadata,
and read the saved copy from then on:

```{r save-extract, eval=FALSE}
extract <- list(
  population = lha_population,
  source = attr(lha_population, "bcdata_source"),
  retrieved = Sys.Date()
)
saveRDS(extract, "data/lha-population-2025-estimates.rds")

# Later, and in every re-run of the report:
extract <- readRDS("data/lha-population-2025-estimates.rds")
lha_population <- extract$population
```

Re-run the retrieval only when the analysis protocol permits the denominator
or geography version to change.

Geography codes alone do not make two periods comparable. Local health area
boundaries are redrawn from time to time, and a code can describe a different
area after a redraw. A historical numerator and a current boundary file need
the same geography vintage, or a documented crosswalk between them.

If a request fails on a health-authority network, verify that
`catalogue.data.gov.bc.ca` is reachable through the configured proxy. The
package does not modify `bcdata` internals or bypass its connectivity checks.

## Common questions

**Why did requesting `sex = "T"` still return F and M rows?** Update
`islandepi` and restart R so the installed namespace matches the current
package. Then inspect `unique(result$sex)`.

**Why are there more population areas than boundaries?** Population resources
cover the province. The default boundary query filters to Island Health.

**Why did a one-to-one join reject the data?** Population output has one row
per age group and sex. Aggregate to one row per geography first, as shown
above, summing over every age band you mean to include.

**How do I reproduce a published map later?** Save the downloaded objects,
their `bcdata_source` attributes and the extraction date in the analysis
project.
