---
title: "Why every estimate needs an interval"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Why every estimate needs an interval}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5,
  fig.retina = 2
)
has_plots <- requireNamespace("islandbrand", quietly = TRUE) &&
  requireNamespace("ggplot2", quietly = TRUE)
```

```{r plot-setup, include=FALSE, eval=has_plots}
islandbrand::islh_setup(format = "plots", quiet = TRUE)
```

```{r packages}
library(dplyr)
library(islandepi)
```

A rate on its own looks exact. "89.5 per 100,000" reads as a precise fact, but
it may rest on two cases or on two hundred. Showing the count helps. The
confidence interval goes further: it shows how far the underlying rate could
plausibly be from the one printed.

This article works through five situations that come up in PHASE products. In
each one, the point estimate alone invites a wrong conclusion, and the
interval guards against it. Every example is evaluated when the website is built,
so the printed values are actual package output.

## Every case is counted. Why an interval?

Surveillance counts every reported case, so the count is not an estimate from a
sample. But the questions a report answers are about the underlying rate: is
it higher in this area than that one? Is it higher this year than last?

The cases that occur in a year are one outcome of that underlying rate. With
exactly the same rate, a different year would bring a different count, in the
same way that a fair coin does not land heads exactly half the time. Treating
the count as a Poisson count, the confidence interval gives the underlying
rates that are compatible with what was observed. This is the standard
approach for routine public health statistics (Eayres 2008).

### What the interval assumes

The interval comes from a model, and it is only as good as the model's
assumptions:

- **Cases are independent Poisson counts.** Each case arrives independently
  of the others.
- **The denominator is fixed and known.** A population estimate is treated as
  exact.

Under those assumptions the interval describes the uncertainty in the rate of
*reported* events. That is not the rate of infection or of disease:
under-reporting, testing practice and case definitions shape the reported
rate, and the interval says nothing about them.

Real counts often vary more than Poisson counts do. Cases in an outbreak are
linked, so they arrive in clusters; one person can have several encounters or
admissions; records from one facility can move together. This extra-Poisson
variation, or overdispersion, makes Poisson intervals too narrow (Spiegelhalter
2005). Counting recurrent events is legitimate, but it does not make Poisson
intervals right for them. When cases are likely to cluster, read the interval
as the least uncertainty there is.

A confidence interval is about the underlying rate, not about next year's
count. Predicting a future count also has to allow for that count's own
chance variation, so a prediction interval is wider.

## 1. The same rate, very different certainty

Island Health's 14 local health areas (LHAs) range from about 2,200 people to
about 259,000. The populations below are BC Stats' 2025 estimates. The case
counts in this article are illustrative.

```{r lha}
lha <- islh_lha_population |>
  summarise(population = sum(population), .by = geography_name) |>
  rename(lha = geography_name) |>
  arrange(desc(population))
lha
```

Suppose the smallest and the largest LHA report the same rate:

```{r same-rate}
same <- lha |>
  filter(lha %in% c("Greater Victoria", "Vancouver Island West")) |>
  mutate(
    cases = c(233, 2),
    islh_crude_rate(cases, population) |> select(rate, lower, upper)
  )

same |>
  mutate(across(c(rate, lower, upper), \(x) round(x, 1)))
```

```{r same-facts, include=FALSE}
victoria <- same |> filter(lha == "Greater Victoria")
west <- same |> filter(lha == "Vancouver Island West")
```

Both rates are about 90 per 100,000. In Greater Victoria the interval runs from
`r round(victoria$lower)` to `r round(victoria$upper)`: from about
`r round(100 * (1 - victoria$lower / victoria$rate))`% below the rate to
`r round(100 * (victoria$upper / victoria$rate - 1))`% above it. In Vancouver
Island West it runs from `r round(west$lower)` to `r round(west$upper)`. One
more case there would add `r round(1 / west$population * 1e5)` per 100,000 to
the rate. Without the interval, a reader cannot tell these two numbers apart.

For a given confidence level and method, the relative width of a rate's
interval, its width compared with the rate itself, depends almost entirely on
the number of events. The population sets the scale: the same count in a
smaller population gives a higher rate and an interval that is wider in rate
units, but no wider relative to the rate. The table shows the relative width
for different counts, using the exact Poisson interval (Garwood 1936):

```{r width-by-count}
tibble(events = c(1, 2, 5, 10, 20, 50, 100, 500)) |>
  mutate(
    islh_ci_poisson(events, method = "exact") |> select(lower, upper),
    below_estimate = sprintf("%.0f%%", 100 * (lower / events - 1)),
    above_estimate = sprintf("+%.0f%%", 100 * (upper / events - 1)),
    across(c(lower, upper), \(x) round(x, 1))
  )
```

With 10 events, the true value could plausibly be half the estimate or nearly
double it. Even at 20 events the interval runs from about 40% below to over
50% above. Rates for small areas, short periods and rare conditions are where
intervals matter most, and those are exactly the rates that tend to make
headlines.

## 2. Ranking areas: the league table trap

A table or map that sorts areas by rate invites the reader to ask what is going
wrong at the top and right at the bottom. When the areas' underlying rates are
similar, small areas tend to fill both ends, because their rates vary most by
chance.

To show this, give every LHA exactly the same underlying rate, 50 per 100,000
per year, and simulate one year of cases:

```{r league-table}
set.seed(61)
true_rate <- 50

league <- lha |>
  mutate(
    cases = stats::rpois(n(), population * true_rate / 1e5),
    islh_crude_rate(cases, population) |> select(rate, lower, upper)
  ) |>
  arrange(desc(rate))

league |>
  mutate(across(c(rate, lower, upper), \(x) round(x, 1)))
```

```{r league-facts, include=FALSE}
top <- league |> slice_head(n = 1)
bottom <- league |> slice_tail(n = 1)
covered <- league |>
  filter(lower <= true_rate, upper >= true_rate) |>
  nrow()
fold <- round(top$rate / bottom$rate)
league_alt <- sprintf(
  paste(
    "Dot-and-whisker chart of crude rates per 100,000 for Island Health's 14",
    "local health areas, sorted from highest to lowest, with a dashed vertical",
    "line at the true rate of %s. %s has the highest rate, about %s, with an",
    "interval from about %s to %s. %s has the lowest, about %s, with an",
    "interval from about %s to %s. %s of the 14 intervals cross the dashed",
    "line, and the intervals for the largest areas are the narrowest."
  ),
  true_rate,
  top$lha, round(top$rate), round(top$lower), round(top$upper),
  bottom$lha, round(bottom$rate), signif(bottom$lower, 1), round(bottom$upper),
  covered
)
```

Sorted by rate, `r top$lha` comes first at `r round(top$rate, 1)` per 100,000,
from `r top$cases` cases. `r bottom$lha` comes last at
`r round(bottom$rate, 1)`, from `r bottom$cases` case. That is a
`r fold`-fold gap between two areas with identical risk, produced by chance
alone.

The intervals tell the real story. `r if (covered == nrow(league)) "All 14" else paste(covered, "of the 14")`
intervals contain the true rate of 50. The two areas at the extremes are among
the smallest, and their intervals are among the widest.

```{r league-plot, eval=has_plots, fig.height=5.5, fig.alt=league_alt}
league |>
  mutate(lha = factor(lha, levels = rev(lha))) |>
  ggplot2::ggplot(ggplot2::aes(x = rate, y = lha)) +
  ggplot2::geom_vline(
    xintercept = true_rate,
    linetype = "dashed",
    colour = islandbrand::islh_hex("grey", 50)
  ) +
  ggplot2::geom_pointrange(
    ggplot2::aes(xmin = lower, xmax = upper),
    colour = islandbrand::islh_brand("primary")
  ) +
  ggplot2::annotate(
    "text",
    x = true_rate,
    y = nrow(league) + 0.9,
    label = "True rate in every area",
    hjust = -0.05,
    size = 3.5
  ) +
  ggplot2::scale_y_discrete(expand = ggplot2::expansion(add = c(0.6, 1.4))) +
  ggplot2::labs(
    x = "Cases per 100,000 (95% confidence interval)",
    y = NULL,
    title = paste0(
      "Same underlying risk, a ",
      fold,
      "-fold range of observed rates"
    ),
    caption = "Simulated cases; populations from BC Stats 2025 estimates."
  )
```

### Use a funnel plot instead of a ranking

A funnel plot shows each area's rate against its population. The funnel marks
where rates would fall by chance alone if every area had the overall rate. It
narrows as the population grows, because larger areas vary less.

An area outside the funnel deserves a closer look. An area inside it does not
stand out from the rest, however high it sits in a ranked list. Draw two
funnels: 95% limits and 99.8% limits (Spiegelhalter 2005). The levels are
nominal. Counts are whole numbers, the limits are interpolated between them,
and the funnel's centre is itself estimated from the same areas, so the
chance of an area crossing them is close to, but not exactly, 1 in 20 and
1 in 500.

This is a **crude-rate Poisson funnel**. Its limits treat each area's count
as Poisson, with expected cases equal to the population times the pooled
crude rate. That suits this simulation, where every area shares one
underlying rate. It does not suit areas whose age structures differ, and
standardized rates cannot simply be put in its place: a standardized rate's
precision depends on each age group's population, rate and weight, not on
the total population, so these limits would be wrong for it. The next
section shows the standard route for age-adjusted comparisons.

Real counts often vary more than Poisson counts, which widens the funnel, and
the overall rate is itself an estimate. Being outside the funnel warrants
review. It does not establish poor performance, and it does not explain why.

Poisson counts are whole numbers, so funnel limits read straight from
`stats::qpois()` jump in steps. Spiegelhalter (2005) interpolates between the
steps to give smooth limits:

```{r funnel-data}
funnel_limit <- function(p, expected) {
  r <- stats::qpois(p, expected)
  step <- (stats::ppois(r, expected) - p) / stats::dpois(r, expected)
  pmax(r - step, 0)
}

overall <- sum(league$cases) / sum(league$population) * 1e5

# Limits for a rate per 100,000, given each population.
rate_limits <- function(population, overall) {
  expected <- population * overall / 1e5
  tibble(
    lower_998 = funnel_limit(0.001, expected),
    lower_95 = funnel_limit(0.025, expected),
    upper_95 = funnel_limit(0.975, expected),
    upper_998 = funnel_limit(0.999, expected)
  ) |>
    mutate(across(everything(), \(x) x / population * 1e5))
}

funnel <- tibble(
  population = exp(seq(log(1500), log(300000), length.out = 300))
) |>
  mutate(rate_limits(population, overall))

league_check <- league |>
  mutate(
    rate_limits(population, overall),
    inside = between(rate, lower_95, upper_95)
  )
sum(league_check$inside)
```

```{r funnel-facts, include=FALSE}
# The plot title and description claim every area is inside the funnel.
stopifnot(all(league_check$inside))
funnel_alt <- sprintf(
  paste(
    "Funnel plot of crude rate per 100,000 against population on a log scale",
    "for the 14 local health areas. A solid horizontal line marks the overall",
    "rate of about %s. Dashed 95%% and dotted 99.8%% limits form a smooth",
    "funnel that is wide for small populations and narrows as population",
    "grows. All 14 areas fall inside the 95%% funnel, including the two",
    "labelled extremes, %s at about %s and %s at about %s."
  ),
  round(overall),
  top$lha, round(top$rate), bottom$lha, round(bottom$rate)
)
limit_colour <- if (has_plots) islandbrand::islh_hex("grey", 45)
```

```{r funnel-plot, eval=has_plots, fig.alt=funnel_alt}
funnel |>
  ggplot2::ggplot(ggplot2::aes(x = population)) +
  ggplot2::geom_line(ggplot2::aes(y = lower_998), linetype = "dotted", colour = limit_colour) +
  ggplot2::geom_line(ggplot2::aes(y = upper_998), linetype = "dotted", colour = limit_colour) +
  ggplot2::geom_line(ggplot2::aes(y = lower_95), linetype = "dashed", colour = limit_colour) +
  ggplot2::geom_line(ggplot2::aes(y = upper_95), linetype = "dashed", colour = limit_colour) +
  ggplot2::geom_hline(yintercept = overall, colour = limit_colour) +
  ggplot2::geom_point(
    data = league,
    ggplot2::aes(y = rate),
    colour = islandbrand::islh_brand("primary"),
    size = 2.5
  ) +
  ggplot2::geom_text(
    data = bind_rows(top, bottom),
    ggplot2::aes(y = rate, label = lha),
    hjust = -0.12,
    size = 3.5
  ) +
  ggplot2::scale_x_log10(
    labels = function(x) format(x, big.mark = ",", scientific = FALSE)
  ) +
  ggplot2::labs(
    x = "Population (log scale)",
    y = "Cases per 100,000",
    title = "Every area sits inside the funnel",
    caption = "Crude rates. Dashed: 95% limits. Dotted: 99.8% limits. Simulated cases."
  )
```

### Areas with different age structures

Most conditions are far more common at older ages, and Island Health's areas
differ in age: some have twice the share of people aged 65 and over that
others have. Crude rates then differ even when the risk at every age is the
same.

The usual route for a funnel is indirect standardization (Spiegelhalter
2005). Apply one set of age-specific reference rates to each area's own
population to get its *expected* count. Plot the ratio of observed to
expected cases against the expected count, and draw the Poisson limits for
the observed count around a ratio of 1. The expected count, not the
population, sets each area's precision.

The simulation below uses the real age structure of every LHA from
`islh_lha_population` and gives every area the *same* rate at each age,
rising steeply with age:

```{r by-age}
age_rates <- islh_lha_population |>
  distinct(age_group) |>
  mutate(
    from_age = as.numeric(sub("[-+].*", "", age_group)),
    rate = 10 * exp(0.05 * from_age)
  )

set.seed(111)
by_age <- islh_lha_population |>
  summarise(population = sum(population), .by = c(geography_name, age_group)) |>
  inner_join(
    age_rates,
    by = join_by(age_group),
    relationship = "many-to-one",
    unmatched = "error"
  ) |>
  mutate(cases = stats::rpois(n(), population * rate / 1e5))

# Reference rates: Island Health's own rate at each age.
reference_rates <- by_age |>
  summarise(reference_rate = sum(cases) / sum(population), .by = age_group)

areas <- by_age |>
  inner_join(reference_rates, by = join_by(age_group)) |>
  # summarise() reads each new column in the expressions after it, so the
  # area's total population comes last.
  summarise(
    aged_65_plus = sum(population[from_age >= 65]) / sum(population),
    observed = sum(cases),
    expected = sum(population * reference_rate),
    population = sum(population),
    .by = geography_name
  ) |>
  mutate(
    crude_rate = observed / population * 1e5,
    ratio = observed / expected
  )

crude_overall <- sum(areas$observed) / sum(areas$population) * 1e5
areas <- areas |>
  mutate(
    crude_outside = !between(
      crude_rate,
      funnel_limit(0.025, population * crude_overall / 1e5) / population * 1e5,
      funnel_limit(0.975, population * crude_overall / 1e5) / population * 1e5
    ),
    ratio_outside = !between(
      ratio,
      funnel_limit(0.025, expected) / expected,
      funnel_limit(0.975, expected) / expected
    )
  )

areas |>
  select(geography_name, aged_65_plus, crude_rate, ratio, crude_outside, ratio_outside) |>
  arrange(desc(aged_65_plus)) |>
  mutate(across(c(aged_65_plus, ratio), \(x) round(x, 2)), crude_rate = round(crude_rate))
```

```{r by-age-facts, include=FALSE}
crude_flagged <- sum(areas$crude_outside)
ratio_flagged <- sum(areas$ratio_outside)
ratio_alt <- sprintf(
  paste(
    "Funnel plot of the ratio of observed to expected cases against the",
    "expected count on a log scale for the 14 local health areas, with a",
    "solid line at a ratio of 1 and dashed 95%% and dotted 99.8%% limits that",
    "narrow as the expected count grows. %s of the 14 areas fall outside the",
    "95%% limits."
  ),
  if (ratio_flagged == 0) "None" else ratio_flagged
)
```

Against a crude-rate funnel, `r crude_flagged` of the 14 areas fall outside
the 95% limits, the older areas above and the younger ones below, although
every area has the same risk at every age. Against the age-standardized
funnel, `r if (ratio_flagged == 0) "none does" else paste(ratio_flagged, "do")`.

```{r ratio-funnel, eval=has_plots, fig.alt=ratio_alt}
ratio_funnel <- tibble(
  expected = exp(seq(
    log(0.8 * min(areas$expected)),
    log(1.25 * max(areas$expected)),
    length.out = 300
  ))
) |>
  mutate(
    lower_998 = funnel_limit(0.001, expected) / expected,
    lower_95 = funnel_limit(0.025, expected) / expected,
    upper_95 = funnel_limit(0.975, expected) / expected,
    upper_998 = funnel_limit(0.999, expected) / expected
  )

ratio_funnel |>
  ggplot2::ggplot(ggplot2::aes(x = expected)) +
  ggplot2::geom_line(ggplot2::aes(y = lower_998), linetype = "dotted", colour = limit_colour) +
  ggplot2::geom_line(ggplot2::aes(y = upper_998), linetype = "dotted", colour = limit_colour) +
  ggplot2::geom_line(ggplot2::aes(y = lower_95), linetype = "dashed", colour = limit_colour) +
  ggplot2::geom_line(ggplot2::aes(y = upper_95), linetype = "dashed", colour = limit_colour) +
  ggplot2::geom_hline(yintercept = 1, colour = limit_colour) +
  ggplot2::geom_point(
    data = areas,
    ggplot2::aes(y = ratio),
    colour = islandbrand::islh_brand("primary"),
    size = 2.5
  ) +
  ggplot2::scale_x_log10() +
  ggplot2::labs(
    x = "Expected cases (log scale)",
    y = "Observed / expected",
    title = "Age-standardized, every area is within the limits",
    caption = paste(
      "Indirectly standardized to Island Health's age-specific rates; simulated cases.",
      "Dashed: 95% limits. Dotted: 99.8% limits.",
      sep = "\n"
    )
  )
```

The ratio of observed to expected cases is an indirectly standardized ratio.
It compares each area with the reference rates, not one area with another.
To compare two areas directly, use directly standardized rates and the ratio
method in the next section.

## 3. Comparing two rates

### Crude rates: "Up 57% on last year"

Suppose the Comox Valley had 14 cases last year and 22 this year, with no
change in population. A headline might read "Cases up 57%".

Compare two rates with a rate ratio and the interval for that ratio.
`islh_rate_ratio()` gives the exact interval for the ratio of two Poisson
rates, conditional on their total (Breslow and Day 1987); it is the interval
`stats::poisson.test()` gives, and it works across the rows of a table. Put
the later period, or the group of interest, first:

```{r small-change}
comox <- tibble(year = c(2024, 2025), cases = c(14, 22), population = 80091)

comox |>
  mutate(
    islh_crude_rate(cases, population) |> select(rate, lower, upper),
    across(c(rate, lower, upper), \(x) round(x, 1))
  )

comox_change <- islh_rate_ratio(
  cases = 22,
  population = 80091,
  reference_cases = 14,
  reference_population = 80091
)
comox_change |>
  select(ratio, lower, upper) |>
  round(2)
```

Reported cases did rise, from 14 to 22, and a report can say so. The rate
ratio, `r round(comox_change$ratio, 2)`, estimates how much the underlying rate
of reported cases changed, and its interval, `r round(comox_change$lower, 2)` to
`r round(comox_change$upper, 2)`, says how precisely. This is an increase
estimated very imprecisely: the counts are compatible with an underlying rate
about a quarter lower than last year's, with one more than three times as
high, and with everything between.

Report the counts, the ratio and its interval, and describe it that way. An
interval that contains 1 does not show that nothing changed, and one that
excludes 1 does not prove a change occurred (Greenland et al. 2016). The
interval describes the size of the change the data support, under the
model.

### Overlapping intervals can still hide a real change

The opposite mistake is just as common. Suppose Greater Nanaimo went from 160
cases to 205:

```{r overlap}
nanaimo <- tibble(year = c(2024, 2025), cases = c(160, 205), population = 133395) |>
  mutate(islh_crude_rate(cases, population) |> select(rate, lower, upper))

nanaimo |>
  mutate(across(c(rate, lower, upper), \(x) round(x, 1)))

nanaimo_change <- islh_rate_ratio(205, 133395, 160, 133395)
nanaimo_change |>
  select(ratio, lower, upper) |>
  round(2)
```

The two intervals overlap: 2024's runs up to `r round(nanaimo$upper[1])`
and 2025's starts at `r round(nanaimo$lower[2])`. It is tempting to read
that overlap as "no change", and it would be wrong. The interval for the
ratio, `r round(nanaimo_change$lower, 2)` to `r round(nanaimo_change$upper, 2)`,
lies above 1. Under the Poisson model, these counts support an increase in the
underlying rate of reported cases, of somewhere between about
`r round(100 * (nanaimo_change$lower - 1))`% and
`r round(100 * (nanaimo_change$upper - 1))`%.

That is evidence the reported rate changed, not of why. A change in testing,
coding or case definitions would produce the same pattern, and clustering of
cases would make the true interval wider than this one.

Two 95% intervals can overlap even when the difference between the estimates is
clear (Schenker and Gentleman 2001). Compare two estimates with the interval
for their ratio or difference, not by checking whether their intervals touch.
For two independent estimates like these, the reverse does hold: when their
95% intervals do not overlap at all, the difference is clear at an even
stricter level than 5%.

### Standardized rates need their own ratio

A directly standardized rate is a weighted sum of age-specific rates. Its
uncertainty comes from each age group's count, population and weight, so it
cannot be recovered from the total cases and total population. Comparing
those totals with `islh_rate_ratio()` or `stats::poisson.test()` compares the
crude rates, whatever the rates on the page say.

`islh_dsr_ratio()` compares two rates from `islh_dsr()` or
`islh_dsr_joined()` with Fay's F interval, built for standardized rates
(Fay 1999). It extends the Fay and Feuer interval those functions give
(Fay and Feuer 1997). Here are two areas from the simulation above, with
the same risk at every age but very different age structures:

```{r dsr-ratio}
standard <- islh_standard_population("canada_2011")

standardized_rate <- function(area) {
  area_by_age <- by_age |> filter(geography_name == area)
  islh_dsr_joined(
    events = area_by_age |> select(age_group, cases),
    population = area_by_age |> select(age_group, population),
    standard = standard,
    by = "age_group",
    standard_id = standard$standard_id[1]
  )
}

oceanside <- standardized_rate("Oceanside")
western <- standardized_rate("Western Communities")

crude_ratio <- islh_rate_ratio(
  oceanside$cases,
  oceanside$population,
  western$cases,
  western$population
)
standardized_ratio <- islh_dsr_ratio(oceanside, western)

bind_rows(crude = crude_ratio, standardized = standardized_ratio, .id = "rates") |>
  select(rates, ratio, lower, upper, method) |>
  mutate(across(c(ratio, lower, upper), \(x) round(x, 2)))
```

```{r dsr-facts, include=FALSE}
share_old <- areas |>
  filter(geography_name %in% c("Oceanside", "Western Communities")) |>
  select(geography_name, aged_65_plus)
```

Oceanside's population is `r round(100 * share_old$aged_65_plus[share_old$geography_name == "Oceanside"])`%
aged 65 and over, the Western Communities' only
`r round(100 * share_old$aged_65_plus[share_old$geography_name == "Western Communities"])`%.
The crude ratio, `r round(crude_ratio$ratio, 2)`, is the difference in age
structure. The standardized ratio, `r round(standardized_ratio$ratio, 2)`
(`r round(standardized_ratio$lower, 2)` to `r round(standardized_ratio$upper, 2)`),
removes it, and its interval is wider because the comparison now rests on
the age groups separately.

Both rates must use the same standard population, and their counts must be
independent: different areas, or one area in two separate periods. An area
and a region that contains it share cases, so they cannot be compared this
way.

## 4. Percentages and targets

Proportions behave the same way. `islh_proportion()` uses the Wilson score
interval (Wilson 1927), which stays between 0 and 100% and, for small
denominators, comes close to the nominal coverage on average. It does not
reach it for every case: with 20 cases and a true proportion of 5%, a nominal
95% Wilson interval covers about 92.5% of the time. `method = "exact"`, the
Clopper-Pearson interval, never falls below the nominal level, and is wider
for it (Brown, Cai and DasGupta 2001). Suppose the target is to reach 90% of
new cases within 24 hours:

```{r proportions}
follow_up <- tibble(
  team = c("Small team", "Large team", "Small team, next quarter"),
  reached = c(18, 1710, 19),
  cases = c(20, 1900, 20)
) |>
  mutate(
    islh_proportion(reached, cases, per = 100) |>
      select(proportion, lower, upper)
  )

follow_up |>
  mutate(across(c(proportion, lower, upper), \(x) round(x, 1)))
```

Both teams reached 90%. The large team's interval is under 3 points wide, so
its performance is known closely. The small team's runs from
`r round(follow_up$lower[1])` to `r round(follow_up$upper[1])`.

The next quarter, the small team reaches 19 of 20, or 95%. As a description
of that quarter, the result exceeded the 90% target, and a report can say so.

Whether the team's ongoing performance now exceeds 90% is a different
question. The interval, `r round(follow_up$lower[3])` to
`r round(follow_up$upper[3])`, is compatible with an underlying rate well
below 90%, so these data do not establish it. One more or one fewer case
reached moves the result by 5 points.

Decide in the reporting policy which of the two statements a target is about.
When it is about ongoing performance, check where the interval sits, not just
the estimate.

## 5. Putting intervals in products

### Tables

Give each estimate its interval in the same cell or the next column. Write the
range with "to" rather than a hyphen, which is easy to misread as a minus sign:

```{r format-table}
format_interval <- function(estimate, lower, upper, digits = 1) {
  fmt <- function(x) formatC(x, format = "f", digits = digits, big.mark = ",")
  sprintf("%s (%s to %s)", fmt(estimate), fmt(lower), fmt(upper))
}

league |>
  arrange(lha) |>
  transmute(
    LHA = lha,
    Cases = cases,
    `Rate per 100,000 (95% CI)` = format_interval(rate, lower, upper)
  ) |>
  slice_head(n = 5)
```

Sort tables of areas alphabetically or geographically, not by rate. Put the
interval method and confidence level in the table notes. `islh_crude_rate()`,
`islh_proportion()` and `islh_dsr()` return them in the `method` and `conf`
columns.

For branded HTML and Word tables, pass the formatted data frame to
`islandbrand::islh_gt()` or `islandbrand::islh_flextable()`.

### Figures

Show intervals as error bars or dot-and-whisker charts for areas and groups,
and as a shaded band for a trend over time. The islandbrand article
[Surveillance figures](https://temuulene.github.io/islandbrand/articles/surveillance.html#showing-uncertainty)
shows both in the Island Health style.

Avoid bar charts of small-area rates without intervals. A bar's length implies
a precision the data do not have.

### Text

State the estimate with its interval at first mention:

> The rate in Greater Victoria was 90.0 cases per 100,000 (95% CI 78.8 to
> 102.3).

For a comparison, report the ratio or difference, its interval and the counts
behind it, and say what the interval shows rather than whether it crosses 1:

> Comox Valley reported 22 cases in 2025, up from 14 in 2024. The rate ratio
> was 1.57 (95% CI 0.77 to 3.32): an increase, estimated too imprecisely to
> say how large it is, or to rule out a fall.

For audiences outside epidemiology, one sentence is enough:

> The range in brackets shows the underlying rates that are consistent with
> the number of cases observed. A wide range means the rate is uncertain.

Avoid saying there is a 95% probability that the true rate lies in the
interval. That is not what a confidence interval means, and the plain-language
sentence above does not need it.

### Standardized rates

Standardized rates need intervals as much as crude rates, and theirs are often
wider, because a few sparse age bands with large weights can dominate. Use
`islh_dsr()` or `islh_dsr_joined()` with the default Fay-Feuer interval,
`islh_dsr_detail()` to see which age bands make the interval wide, and
`islh_dsr_ratio()` to compare two of them. The
[Getting started](islandepi.html#directly-standardized-rates) article has a
worked example.

The Fay-Feuer interval is conservative by design: it usually covers more
often than its nominal level. Tools built on Dobson's method, such as the
Fingertips profiles, give somewhat narrower intervals, so the two will not
match exactly (Eayres 2008).

## What an interval does not cover

A confidence interval describes chance variation in the count, under the
model. It does not cover:

- under-reporting, or changes in testing, case definitions or coding;
- cases not yet reported for recent periods, which
  `islh_reporting_completeness()` estimates;
- errors in the population denominator;
- extra variation from clustered or recurrent events, beyond what a Poisson
  count has;
- the extra chance findings that come from making many comparisons at once.

On the last point: compare each of 14 areas with the overall rate at the 95%
level, and about 0.7 areas are expected to look unusual by chance alone in a
year, with about a 51% chance that at least one does. That assumes the 14
comparisons are independent and each has exactly a 5% chance of a false
alarm. With whole-number counts the actual chances differ a little, and
comparisons against an overall rate that includes the area are not quite
independent.

When these matter, say so beside the interval rather than leaving the reader
to assume the interval covers them.

## Checklist for PHASE products

- Give every rate, proportion and standardized rate its 95% interval, in
  tables, figures and text.
- Compare two estimates with the interval for their ratio or difference, not by
  eye and not by whether their intervals overlap. Use `islh_rate_ratio()` for
  crude rates and `islh_dsr_ratio()` for standardized rates.
- Describe the estimate, its interval and the model behind it. Do not turn the
  interval into a yes-or-no test of whether it crosses 1.
- Do not rank small areas by their point estimates. Use a funnel plot, or sort
  by name or geography and show the intervals. Standardize the funnel when
  age structures differ.
- Before saying a result meets, exceeds or misses a target, be clear whether
  the statement is about the period observed or about ongoing performance.
  For ongoing performance, check where the interval sits.
- Name the interval method and confidence level in the notes.
- When an interval is too wide to support any conclusion, say so. Consider
  combining years or areas, applied the same way to every estimate being
  compared.
- Keep precision separate from suppression. A count can be large enough to
  publish and still too imprecise to compare.

## References

Breslow NE, Day NE (1987). *Statistical Methods in Cancer Research, Volume
II: The Design and Analysis of Cohort Studies*. IARC Scientific Publications
No. 82. Lyon: International Agency for Research on Cancer.

Brown LD, Cai TT, DasGupta A (2001). Interval estimation for a binomial
proportion. *Statistical Science* 16(2):101-133.
<https://doi.org/10.1214/ss/1009213286>

Eayres D (2008). *Technical Briefing 3: Commonly used public health statistics
and their confidence intervals*. York: Association of Public Health
Observatories.

Fay MP (1999). Approximate confidence intervals for rate ratios from directly
standardized rates with sparse data. *Communications in Statistics - Theory
and Methods* 28(9):2141-2160.

Fay MP, Feuer EJ (1997). Confidence intervals for directly standardized rates:
a method based on the gamma distribution. *Statistics in Medicine*
16(7):791-801.

Garwood F (1936). Fiducial limits for the Poisson distribution. *Biometrika*
28(3-4):437-442.

Greenland S, Senn SJ, Rothman KJ, Carlin JB, Poole C, Goodman SN, Altman DG
(2016). Statistical tests, P values, confidence intervals, and power: a guide
to misinterpretations. *European Journal of Epidemiology* 31(4):337-350.
<https://doi.org/10.1007/s10654-016-0149-3>

Schenker N, Gentleman JF (2001). On judging the significance of differences by
examining the overlap between confidence intervals. *The American
Statistician* 55(3):182-186.

Spiegelhalter DJ (2005). Funnel plots for comparing institutional
performance. *Statistics in Medicine* 24(8):1185-1202.
<https://doi.org/10.1002/sim.1970>

Wilson EB (1927). Probable inference, the law of succession, and statistical
inference. *Journal of the American Statistical Association*
22(158):209-212.
