---
title: "Getting started with islandepi"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with islandepi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(islandepi)
```

`islandepi` provides small, validated functions for routine public health
analysis. Every example on this page is evaluated when the website is built,
so the printed values are actual package output.

The package implements calculation mechanics. It does not choose the
suppression threshold, standard population or interval method for a release.

Every rate, proportion and standardized rate below comes with a confidence
interval. Report the interval with the estimate. The article
[Why every estimate needs an interval](intervals.html) shows, with worked
examples, how leaving it out leads to wrong conclusions.

## Choose the function

| Task | Function | Output |
|---|---|---|
| Group individual ages | `islh_age_group()` | Ordered factor |
| Attach denominators with checked keys | `islh_join_denominator()` | Counts with a denominator column |
| Calculate a crude rate | `islh_crude_rate()` | Data frame with rate and interval |
| Calculate a proportion or percentage | `islh_proportion()` | Data frame with estimate and interval |
| Calculate a Poisson count interval | `islh_ci_poisson()` | Data frame with lower and upper limits |
| Calculate a directly standardized rate | `islh_dsr_joined()`, `islh_dsr()` | One-row data frame |
| Get a published standard population | `islh_standard_population()` | Age bands with weights |
| See which strata drive a standardized rate | `islh_dsr_detail()` | One row per stratum |
| Hide small counts | `islh_suppress()` | Numeric or labelled character vector |
| Apply suppression to a table | `islh_suppress_table()` | Updated data frame |
| Round counts | `islh_round_base()` | Numeric vector |
| Compare with an existing report | `islh_compare_outputs()` | Differences, one row per cell |
| Retrieve BC denominators | `islh_bc_population()` | Tidy population table |
| Retrieve BC boundaries | `islh_bc_geography()` | Standardized `sf` object |

## Example data

The examples use `islh_outbreak`, a simulated line list bundled with the
package: 790 cases of a winter respiratory virus across Island Health's local
health areas, November 2025 to March 2026. It was simulated with the
[simulist](https://epiverse-trace.github.io/simulist/) package. No record
describes a real person.

```{r example-data}
head(islh_outbreak[c("case_id", "age", "sex", "date_onset", "hsda", "lha_name")])
```

## Group ages

The result is an ordered factor, which preserves the correct order in tables
and plots.

```{r age-groups}
cases <- islh_outbreak
cases$age_group <- islh_age_group(cases$age, breaks = "broad")
table(cases$age_group)
```

Built-in choices are `"five_year"`, `"phac"` and `"broad"`. A local set of
lower bounds also works:

```{r custom-age-groups}
islh_age_group(c(2, 17, 42, 91), breaks = c(0, 18, 65))
```

Missing ages become `NA`. Negative ages also become `NA`, with a warning,
because they usually point to a wrong birth or event date. Check those rows
before calculating a rate rather than silently dropping them.

## Crude rates

`islh_crude_rate()` returns the numerator and denominator beside the estimate
and confidence limits, making the calculation easy to audit.

The case counts come from the simulated outbreak. The denominators are BC
Stats' 2025 population estimates for each health service delivery area.

Keep counts and denominators in separate tables and join them on explicit
keys. `islh_join_denominator()` stops if a count has no denominator, if the
denominator table has more than one row for a key, or if a key is text in one
table and a number in the other:

```{r crude-rate}
hsda_population <- data.frame(
  hsda = c(
    "South Vancouver Island",
    "Central Vancouver Island",
    "North Vancouver Island"
  ),
  population = c(464081, 320319, 143305)
)
hsda_cases <- as.data.frame(table(hsda = cases$hsda), responseName = "cases")
hsda_cases$hsda <- as.character(hsda_cases$hsda)

by_hsda <- islh_join_denominator(hsda_cases, hsda_population, by = "hsda")

crude <- islh_crude_rate(
  cases = by_hsda$cases,
  population = by_hsda$population,
  per = 100000
)
cbind(
  by_hsda["hsda"],
  round(crude[c("cases", "population", "rate", "lower", "upper")], 1)
)
```

Read each row as that area's cases in its population, expressed per 100,000
people, with the chosen Poisson interval. Counts may exceed the denominator
when the numerator contains repeat events and the denominator is person-time.

`method` records which interval each row actually used. The default,
`method = "auto"`, takes the exact interval for counts below 10 and Byar's
approximation from 10 upwards. Set `method` explicitly when a release has to
match a published series that used one method throughout. The result also
carries `conf` and `per`, so a saved table records how it was calculated.

## Proportions and percentages

Use `islh_proportion()` when the outcome is a share of an eligible group, such
as immunization coverage, test positivity or the share of cases admitted to
hospital. The numerator must be a subset of the denominator.

```{r proportion}
admitted <- tapply(!is.na(cases$date_admission), cases$hsda, sum)
total <- tapply(cases$case_id, cases$hsda, length)

share <- islh_proportion(admitted, total, per = 100)
cbind(
  hsda = names(total),
  round(share[c("x", "n", "proportion", "lower", "upper")], 1)
)
```

The default Wilson score interval stays within 0 to 100% and keeps close to
its stated coverage for small denominators. Use `method = "exact"` for the
Clopper-Pearson interval when a series requires it. Decide how records with
an unknown status are handled before calculating, and say so in the table
notes. Survey-weighted prevalence needs a method that accounts for the survey
design, not these intervals.

## Poisson count intervals

```{r poisson-intervals}
islh_ci_poisson(c(0, 3, 25))
```

The default picks a method per count: exact below 10, Byar at 10 and above.
Set `method = "exact"` or `method = "byar"` to use one method throughout, as
an analytical standard may require. `"byar"` is an approximation and falls
back to the exact interval when the count is zero, because Byar's is undefined
there.

## Directly standardized rates

Keep cases, population and the standard population in three tables keyed by
age group. `islh_dsr_joined()` lines them up by key and stops unless all three
cover exactly the same strata, so a table in a different row order, or one
missing an age band, cannot give a wrong rate.

`islh_standard_population()` returns a published standard in the bands you
ask for. There is no default: use the standard your comparison series uses.
The case and population figures below are illustrative.

```{r standardized-rate}
standard <- islh_standard_population("canada_2011", age_breaks = c(0, 20, 45, 65))
standard$age_group <- as.character(standard$age_group)
standard[c("age_group", "population", "weight")]

age_cases <- data.frame(
  age_group = c("0-19", "20-44", "45-64", "65+"),
  cases = c(5, 12, 40, 80)
)
age_population <- data.frame(
  age_group = c("65+", "45-64", "20-44", "0-19"),
  population = c(15000, 22000, 25000, 20000)
)

dsr <- islh_dsr_joined(
  age_cases,
  age_population,
  standard,
  by = "age_group",
  standard_id = standard$standard_id[1]
)
round(dsr[c("cases", "population", "rate", "lower", "upper")], 1)
```

The returned `population` is the total observed denominator. The standard
population is converted to weights and is not added to the result.
`standard_id` records which standard was used. `method = "gamma"` uses the
Fay-Feuer interval; `"normal"` is available for comparison with analyses that
explicitly require it. `islh_dsr()` takes the same inputs as vectors already
in stratum order.

`islh_dsr_detail()` shows how each stratum feeds the result. The
contributions add up to the standardized rate, and `variance_share` shows which
strata make the interval wide:

```{r dsr-detail}
detail <- islh_dsr_detail(dsr)
detail$rate <- round(detail$rate, 1)
detail$contribution <- round(detail$contribution, 1)
detail$variance_share <- round(detail$variance_share, 2)
detail
```

Here the two oldest bands carry more than 80% of the variance, because their
rates are highest and their populations are smaller. In a small area, one or
two sparse strata can dominate like this. Broader bands, a longer period or a
larger area help, applied the same way to every rate being compared.

Standardization answers one question: would two populations differ if they
had the same age structure? In the example below, both areas have exactly the
same rate in every age band. Area B is older, so its crude rate is higher, but
the standardized rates are identical:

```{r same-risk}
same_risk <- data.frame(
  age_group = c("0-19", "20-44", "45-64", "65+"),
  rate_per_person = c(0.0002, 0.0005, 0.0020, 0.0060)
)
area_a <- data.frame(
  age_group = same_risk$age_group,
  population = c(30000, 40000, 20000, 10000)
)
area_b <- data.frame(
  age_group = same_risk$age_group,
  population = c(15000, 26000, 30000, 30000)
)
area_a$cases <- round(area_a$population * same_risk$rate_per_person)
area_b$cases <- round(area_b$population * same_risk$rate_per_person)

rates <- rbind(
  area_a = c(
    crude = sum(area_a$cases) / sum(area_a$population) * 1e5,
    standardized = islh_dsr_joined(
      area_a[c("age_group", "cases")],
      area_a[c("age_group", "population")],
      standard,
      by = "age_group",
      standard_id = standard$standard_id[1]
    )$rate
  ),
  area_b = c(
    crude = sum(area_b$cases) / sum(area_b$population) * 1e5,
    standardized = islh_dsr_joined(
      area_b[c("age_group", "cases")],
      area_b[c("age_group", "population")],
      standard,
      by = "age_group",
      standard_id = standard$standard_id[1]
    )$rate
  )
)
round(rates, 1)
```

A crude rate is still the right number for planning services, because it
describes the burden in the population as it is. Report both, and say which
one a comparison uses.

## Suppress small cells

There is deliberately no default threshold. Supply the approved rule for the
specific release.

```{r suppress-vector}
counts <- c(0, 3, 5, 17, 42)
islh_suppress(counts, threshold = 5, label = "Suppressed")
```

Zero is kept because it is an observed absence, not a small positive count.
With the default `inclusive = TRUE`, both 3 and 5 are suppressed.

For a table, count the simulated cases and admissions in each North Island
local health area. The small areas have few cases, and fewer admissions:

```{r suppress-table}
north <- cases[cases$hsda == "North Vancouver Island", ]
area_counts <- data.frame(
  area = sort(unique(north$lha_name)),
  cases = as.vector(table(north$lha_name)),
  admitted = as.vector(tapply(
    !is.na(north$date_admission),
    north$lha_name,
    sum
  ))
)
area_counts

islh_suppress_table(
  area_counts,
  cols = c("cases", "admitted"),
  threshold = 5,
  label = "Suppressed"
)
```

A rate or confidence limit calculated from a hidden count gives the count
away: multiply the rate by the population and it is back. `linked` hides
those columns in the same rows:

```{r linked-suppression}
area_rates <- area_counts
area_rates$rate <- round(area_rates$admitted / area_rates$cases * 100, 1)

islh_suppress_table(
  area_rates,
  cols = "admitted",
  threshold = 5,
  linked = list(admitted = "rate")
)
```

Linked cells become `NA` and keep the column numeric. Set `linked_label` to
show a neutral label instead. Either way, the audit records them with the
reason `"linked"`.

Complementary suppression can hide a second cell when one hidden value could
be recovered from a visible total:

```{r complementary-suppression}
islh_suppress_table(
  area_counts,
  cols = "cases",
  threshold = 5,
  label = "Suppressed",
  complementary = TRUE
)
```

By default the complementary rule looks at the whole column, which is right
for a table carrying one overall total. Where a table is subtotalled, name the
subtotal columns with `by` so the rule runs inside each block:

```{r grouped-suppression}
by_authority <- data.frame(
  authority = c("Island", "Island", "Interior", "Interior"),
  area = c("North", "South", "East", "West"),
  cases = c(2, 30, 4, 25)
)

hidden <- islh_suppress_table(
  by_authority,
  cols = "cases",
  threshold = 5,
  label = "Suppressed",
  complementary = TRUE,
  by = "authority"
)

hidden
```

Column-wide, two cells are small and the arithmetic already fails to close.
Within each authority there is exactly one hidden cell, which a reader recovers
by subtracting from that authority's subtotal, so a second cell goes in each.

`islh_suppression_audit()` reports what was hidden and why, so a reviewer can
confirm the rule fired where it should have:

```{r suppression-audit}
islh_suppression_audit(hidden)
```

Keep the audit internal. It records no counts, but the reasons still say
something about the hidden values: a cell marked `"small"` is known to be at or
below the threshold. Publishing the audit, or a table that still carries it,
narrows what a reader can infer. Save the audit with your working notes, then
remove it from the object you release:

```{r release-copy}
audit <- islh_suppression_audit(hidden)
released <- hidden
attr(released, "islh_suppression") <- NULL
names(attributes(released))
```

This helper works independently within each selected column. It does not prove
that a multidimensional or linked release is protected. Use `linked` for
rates, percentages and confidence limits calculated from a suppressed count.

Keep suppression separate from missing data and from precision. Without a
`label`, a suppressed cell and a missing value are both `NA`, so mark
suppressed cells in the table notes using the audit. A wide confidence
interval is a separate flag again: a count can be large enough to publish and
still too imprecise to compare.

## Round counts

```{r deterministic-rounding}
islh_round_base(c(0, 2, 3, 7, 12, 43), base = 5)
```

Deterministic rounding returns the same values each time. Random rounding is
unbiased over repeated use but varies between runs. Set a seed, round once and
save the released values:

```{r random-rounding}
set.seed(42)
islh_round_base(c(2, 3, 7, 12), base = 5, method = "random")
```

## Compare with an existing report

Before a new R output replaces a legacy report or a dashboard, check that the
two agree. `islh_compare_outputs()` lines the tables up on their keys and
returns one row for every value that differs, plus rows only one table has.
No rows means full agreement.

```{r compare-outputs}
dashboard <- data.frame(
  hsda = by_hsda$hsda,
  cases = by_hsda$cases,
  rate = round(crude$rate, 1)
)
# Pretend the dashboard counted one case differently.
dashboard$cases[1] <- dashboard$cases[1] + 1

r_output <- data.frame(
  hsda = by_hsda$hsda,
  cases = by_hsda$cases,
  rate = crude$rate
)

islh_compare_outputs(r_output, dashboard, by = "hsda", tolerance = 0.05)
```

`tolerance` absorbs rounding in the reference table's rates. It applies to
every numeric column, so keep it below 1 when counts are compared too, or
compare counts and rates in separate calls.

## Population and geography data

The BC Data Catalogue helpers require a network request and are documented in
[BC population and geography data](bcdata.html). That article shows the exact
returned columns, representative output, a checked join and troubleshooting
for unmatched rows.

Use `islandbrand` after the calculation when the result needs an Island Health
figure, table or Quarto report.
