---
title: "Getting started with islandepi"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with islandepi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

`islandepi` provides small, validated functions for routine public health
analysis. Every example on this page is evaluated when the website is built,
so the printed values are actual package output. The examples use dplyr and
tidyr, which `islandepi` installs with itself.

```{r packages}
library(dplyr)
library(tidyr)
library(islandepi)
```

The package implements calculation mechanics. It does not choose the
suppression threshold, standard population or interval method for a release.

Every rate, proportion and standardized rate below comes with a confidence
interval. Report the interval with the estimate. The article
[Why every estimate needs an interval](intervals.html) shows, with worked
examples, how leaving it out leads to wrong conclusions.

## Choose the function

| Task | Function | Output |
|---|---|---|
| Group individual ages | `islh_age_group()` | Ordered factor |
| Attach denominators with checked keys | `islh_join_denominator()` | Counts with a denominator column |
| Calculate a crude rate | `islh_crude_rate()` | Data frame with rate and interval |
| Compare two crude rates | `islh_rate_ratio()` | Ratio with an exact interval |
| Calculate a proportion or percentage | `islh_proportion()` | Data frame with estimate and interval |
| Calculate a Poisson count interval | `islh_ci_poisson()` | Data frame with lower and upper limits |
| Calculate a directly standardized rate | `islh_dsr_joined()`, `islh_dsr()` | One-row data frame |
| Compare two standardized rates | `islh_dsr_ratio()` | Ratio with Fay's F interval |
| Get a published standard population | `islh_standard_population()` | Age bands with weights |
| See which strata drive a standardized rate | `islh_dsr_detail()` | One row per stratum |
| Hide small counts | `islh_suppress()` | Numeric or labelled character vector |
| Apply suppression to a table | `islh_suppress_table()` | Updated data frame |
| Round counts | `islh_round_base()` | Numeric vector |
| Compare with an existing report | `islh_compare_outputs()` | Differences, one row per cell |
| Retrieve BC denominators | `islh_bc_population()` | Tidy population table |
| Retrieve BC boundaries | `islh_bc_geography()` | Standardized `sf` object |

## Example data

The examples use `islh_outbreak`, a simulated line list bundled with the
package: 790 cases of a winter respiratory virus across Island Health's local
health areas, November 2025 to March 2026. It was simulated with the
[simulist](https://epiverse-trace.github.io/simulist/) package. No record
describes a real person.

```{r example-data}
islh_outbreak |>
  select(case_id, age, sex, date_onset, hsda, lha_name) |>
  slice_head(n = 6)
```

The denominators are real: `islh_lha_population` holds BC Stats' estimates of
each local health area's population on 1 July 2025, by sex and five-year age
group, as `islh_bc_population()` returns them.

```{r population-data}
islh_lha_population |>
  slice_head(n = 4)
```

## A complete rate workflow

The sections after this one take each function on its own. This one runs the
whole path from line list to a table you could release, so the decisions a
rate depends on are visible in one place: what is counted, over which dates,
against which denominator and standard, with which interval, and what leaves
the team.

The indicator is hospital admissions with the simulated winter virus, by
health service delivery area (HSDA), for onsets from 3 November 2025 to
8 March 2026, age-standardized to the 2011 Canadian population.

That window is 18 weeks. The rates below are admissions per 100,000 people
over those 18 weeks, not per year: they are not annualized. Say so beside
every rate, because a reader will otherwise assume an annual rate.

```{r workflow-counts}
hsdas <- c(
  "South Vancouver Island",
  "Central Vancouver Island",
  "North Vancouver Island"
)
period <- "3 November 2025 to 8 March 2026 (18 weeks)"

# Count every area and age band, including any band with no admissions.
admissions <- islh_outbreak |>
  filter(
    !is.na(date_admission),
    between(date_onset, as.Date("2025-11-03"), as.Date("2026-03-08"))
  ) |>
  mutate(age_group = islh_age_group(age, breaks = "broad")) |>
  count(hsda, age_group, name = "cases") |>
  complete(hsda = hsdas, age_group, fill = list(cases = 0L))

admissions
```

A stratum with no admissions still belongs in a standardized rate. Dropping
it would change the weights. `complete()` adds it back as a zero, because
`age_group` is a factor that knows every band; `islh_count_events()` and its
`groups` argument do the same for counts over time.

The denominators come from `islh_lha_population`, grouped into the same three
bands. In a real analysis, retrieve current estimates with
`islh_bc_population()` and save them with their provenance: see the BC data
article. The standard is the 2011 Canadian population in the same bands.

```{r workflow-denominators}
# The first age in each five-year group, such as 65 for "65-69".
band_start <- function(age_group) as.numeric(sub("[-+].*", "", age_group))

hsda_age_population <- islh_lha_population |>
  mutate(age_group = islh_age_group(band_start(age_group), breaks = "broad")) |>
  summarise(population = sum(population), .by = c(hsda, age_group))

standard <- islh_standard_population("canada_2011", age_breaks = c(0, 20, 65))

hsda_age_population |>
  pivot_wider(names_from = age_group, values_from = population)
```

Each area's rate comes from `islh_dsr_joined()`, which lines the three tables
up by key and stops if any of them is missing an age band:

```{r workflow-rates}
hsda_rate <- function(area) {
  islh_dsr_joined(
    admissions |> filter(hsda == area) |> select(age_group, cases),
    hsda_age_population |> filter(hsda == area) |> select(age_group, population),
    standard,
    by = "age_group",
    standard_id = standard$standard_id[1]
  )
}

rates <- tibble(hsda = hsdas) |>
  reframe(
    hsda_rate(hsda) |> select(cases, population, rate, lower, upper),
    .by = hsda
  ) |>
  mutate(across(c(rate, lower, upper), \(x) round(x, 1)), period = period)

rates
```

```{r workflow-facts, include=FALSE}
widest <- rates |> slice_max(upper - lower, n = 1)
north_south <- islh_dsr_ratio(
  hsda_rate("North Vancouver Island"),
  hsda_rate("South Vancouver Island")
)
```

Every rate carries its 95% Fay-Feuer interval. With `r widest$cases`
admissions, `r widest$hsda`'s interval runs from `r widest$lower` to
`r widest$upper` per 100,000.

Before saying one area's rate is higher than another's, compare them with a
ratio and its interval, not by their order in the table. These are
standardized rates, so use `islh_dsr_ratio()`, which is built for them.
Comparing their total admissions and populations with `islh_rate_ratio()`
or `stats::poisson.test()` would compare the crude rates instead:

```{r workflow-compare}
islh_dsr_ratio(
  hsda_rate("North Vancouver Island"),
  hsda_rate("South Vancouver Island")
) |>
  select(ratio, lower, upper, method)
```

The North Island's rate is estimated at
`r round(north_south$ratio, 1)` times the South Island's, but the interval,
`r round(north_south$lower, 2)` to `r round(north_south$upper, 1)`, is
compatible with the North's rate being half the South's or nearly four times
it. Nine admissions cannot say more. The
[intervals article](intervals.html#comparing-two-rates) explains the
methods.

### Check the age bands

Three broad bands keep the example small, and they are coarse. Each band
still holds a wide range of ages: admission risk rises steeply from 65 to 90,
so an area with more people in their 80s has more admissions in the 65+ band
at the same risk at each age. Using the same broad bands for every area does
not remove that difference. Standardization adjusts only for the age
differences *between* bands.

So the bands are a trade-off. Finer bands adjust more completely; broader
bands give fewer empty or sparse strata and narrower intervals. For a
production indicator, use the age bands of the series you compare with, and
check how much the result moves when bands are pooled:

```{r band-sensitivity}
rates_with <- function(breaks) {
  counts <- islh_outbreak |>
    filter(
      !is.na(date_admission),
      between(date_onset, as.Date("2025-11-03"), as.Date("2026-03-08"))
    ) |>
    mutate(age_group = islh_age_group(age, breaks = breaks)) |>
    count(hsda, age_group, name = "cases") |>
    complete(hsda = hsdas, age_group, fill = list(cases = 0L))
  population <- islh_lha_population |>
    mutate(age_group = islh_age_group(band_start(age_group), breaks = breaks)) |>
    summarise(population = sum(population), .by = c(hsda, age_group))
  weights <- islh_standard_population("canada_2011", age_breaks = breaks)
  rate_for <- function(area) {
    islh_dsr_joined(
      counts |> filter(hsda == area) |> select(age_group, cases),
      population |> filter(hsda == area) |> select(age_group, population),
      weights,
      by = "age_group",
      standard_id = weights$standard_id[1]
    )
  }

  tibble(hsda = hsdas) |>
    reframe(rate_for(hsda) |> select(rate, lower, upper), .by = hsda)
}

bind_rows(
  `three bands` = rates_with(c(0, 20, 65)),
  `five bands` = rates_with(c(0, 20, 45, 65, 80)),
  .id = "age_bands"
) |>
  mutate(across(c(rate, lower, upper), \(x) round(x, 2))) |>
  arrange(hsda)
```

Here five bands, which split off the 80+ group, move each rate by less than
0.1 per 100,000, so the three broad bands are adequate for this comparison.
That will not always hold. Check it whenever areas differ in how old their
oldest residents are, and report the bands used.

Last, apply the release rule and make the copy that leaves the team. The
threshold here is illustrative; use the approved rule for the release:

```{r workflow-release}
checked <- islh_suppress_table(
  rates,
  "cases",
  threshold = 5,
  linked = list(cases = c("rate", "lower", "upper"))
)
released <- islh_release_copy(
  checked,
  c("hsda", "period", "cases", "rate", "lower", "upper")
)
released
```

`checked` is the working object, with its suppression audit for reviewers.
`released` is the plain table to share. It keeps the period, so the rates
cannot be mistaken for annual ones.

## Group ages

The result is an ordered factor, which preserves the correct order in tables
and plots.

```{r age-groups}
islh_outbreak |>
  mutate(age_group = islh_age_group(age, breaks = "broad")) |>
  count(age_group)
```

Built-in choices are `"five_year"`, `"phac"` and `"broad"`. A local set of
lower bounds also works:

```{r custom-age-groups}
islh_age_group(c(2, 17, 42, 91), breaks = c(0, 18, 65))
```

Missing ages become `NA`. Negative ages also become `NA`, with a warning,
because they usually point to a wrong birth or event date. Check those rows
before calculating a rate rather than silently dropping them.

## Crude rates

`islh_crude_rate()` returns the numerator and denominator beside the estimate
and confidence limits, making the calculation easy to audit.

The case counts are every case in the simulated outbreak, onsets from
3 November 2025 to 6 March 2026. The denominators are BC Stats' 2025
population estimates for each health service delivery area, so the rates are
cases per 100,000 people over the whole wave, not per year.

Keep counts and denominators in separate tables and join them on explicit
keys. `islh_join_denominator()` stops if a count has no denominator, if the
denominator table has more than one row for a key, or if a key is text in one
table and a number in the other:

```{r crude-rate}
hsda_population <- islh_lha_population |>
  summarise(population = sum(population), .by = hsda)

hsda_cases <- islh_outbreak |>
  count(hsda, name = "cases")

crude <- islh_join_denominator(hsda_cases, hsda_population, by = "hsda") |>
  mutate(
    islh_crude_rate(cases, population, per = 100000) |>
      select(rate, lower, upper, method)
  )

crude |>
  mutate(across(c(rate, lower, upper), \(x) round(x, 1)))
```

Read each row as that area's cases in its population, expressed per 100,000
people, with the chosen Poisson interval. Counts may exceed the denominator
when the numerator contains repeat events and the denominator is person-time.

The interval treats the count as Poisson and the population as fixed. It
describes the uncertainty in the rate of *reported* cases under that model,
not the rate of infection, and it is too narrow when cases cluster, as they do
in an outbreak. The [intervals article](intervals.html#what-the-interval-assumes)
explains these assumptions.

`method` records which interval each row actually used. The default,
`method = "auto"`, takes the exact interval for counts below 10 and Byar's
approximation from 10 upwards. Set `method` explicitly when a release has to
match a published series that used one method throughout. The result also
carries `conf` and `per`, so a saved table records how it was calculated.

## Proportions and percentages

Use `islh_proportion()` when the outcome is a share of an eligible group, such
as immunization coverage, test positivity or the share of cases admitted to
hospital. The numerator must be a subset of the denominator.

```{r proportion}
islh_outbreak |>
  summarise(
    admitted = sum(!is.na(date_admission)),
    cases = n(),
    .by = hsda
  ) |>
  mutate(
    islh_proportion(admitted, cases, per = 100) |>
      select(proportion, lower, upper),
    across(c(proportion, lower, upper), \(x) round(x, 1))
  )
```

The default Wilson score interval stays within 0 to 100%, and for small
denominators its coverage is close to the stated level on average. It is not
guaranteed for every case: a nominal 95% Wilson interval for 20 people and a
true proportion of 5% covers about 92.5% of the time (Brown, Cai and DasGupta
2001). Use `method = "exact"` for the Clopper-Pearson interval, which never
covers less than the stated level and is wider for it, or when a series
requires it. Decide how records with an unknown status are handled before
calculating, and say so in the table notes. Survey-weighted prevalence needs a
method that accounts for the survey design, not these intervals.

## Poisson count intervals

```{r poisson-intervals}
islh_ci_poisson(c(0, 3, 25))
```

The default picks a method per count: exact below 10, Byar at 10 and above.
Set `method = "exact"` or `method = "byar"` to use one method throughout, as
an analytical standard may require. `"byar"` is an approximation and falls
back to the exact interval when the count is zero, because Byar's is undefined
there.

## Directly standardized rates

Keep cases, population and the standard population in three tables keyed by
age group. `islh_dsr_joined()` lines them up by key and stops unless all three
cover exactly the same strata, so a table in a different row order, or one
missing an age band, cannot give a wrong rate.

`islh_standard_population()` returns a published standard in the bands you
ask for. There is no default: use the standard your comparison series uses.
The case and population figures below are illustrative.

```{r standardized-rate}
standard <- islh_standard_population("canada_2011", age_breaks = c(0, 20, 45, 65))
standard |>
  select(age_group, population, weight)

age_cases <- tibble(
  age_group = c("0-19", "20-44", "45-64", "65+"),
  cases = c(5, 12, 40, 80)
)
age_population <- tibble(
  age_group = c("65+", "45-64", "20-44", "0-19"),
  population = c(15000, 22000, 25000, 20000)
)

dsr <- islh_dsr_joined(
  age_cases,
  age_population,
  standard |> mutate(age_group = as.character(age_group)),
  by = "age_group",
  standard_id = standard$standard_id[1]
)
dsr |>
  select(cases, population, rate, lower, upper) |>
  round(1)
```

The returned `population` is the total observed denominator. The standard
population is converted to weights and is not added to the result.
`standard_id` records which standard was used. `method = "gamma"` uses the
Fay-Feuer interval (Fay and Feuer 1997); `"normal"` is available for
comparison with analyses that explicitly require it. `islh_dsr()` takes the
same inputs as vectors already in stratum order.

The Fay-Feuer interval is conservative by design: it usually covers more often
than its stated level. Tools built on Dobson's method, such as the Fingertips
profiles, give somewhat narrower intervals, so the two will not match exactly.

`islh_dsr_detail()` shows how each stratum feeds the result. The
contributions add up to the standardized rate, and `variance_share` shows each
stratum's share of the estimated variance:

```{r dsr-detail}
islh_dsr_detail(dsr) |>
  mutate(
    across(c(rate, contribution), \(x) round(x, 1)),
    across(c(weight, variance_share), \(x) round(x, 2))
  )
```

Here the two oldest bands carry more than 80% of the variance, because their
rates are highest and their populations are smaller.

The variance is not the whole story of the interval's width. Fay and Feuer's
upper limit also adds the largest weight per person of any stratum. A stratum
with no cases adds nothing to the variance, so its `variance_share` is 0, yet
if its population is small and its weight large it can widen the upper limit a
great deal. Below, three equally weighted strata hold 0, 30 and 60 cases; only
the size of the empty stratum changes:

```{r empty-stratum}
tibble(empty_stratum_population = c(20000, 2000, 200)) |>
  reframe(
    islh_dsr(
      cases = c(0, 30, 60),
      population = c(empty_stratum_population, 20000, 20000),
      std_population = c(1, 1, 1)
    ) |>
      select(rate, lower, upper),
    .by = empty_stratum_population
  ) |>
  mutate(across(c(rate, lower, upper), \(x) round(x, 1)))
```

The rate and the lower limit do not move, while the upper limit nearly
quadruples. So read `weight` and `population` in the detail as well as
`variance_share`. A longer period or a larger area helps; broader age bands
also help, at the cost described in the workflow above. Apply any such choice
the same way to every rate being compared.

Standardization answers one question: would two populations differ if they
had the same age structure? In the example below, both areas have exactly the
same rate in every age band. Area B is older, so its crude rate is higher, but
the standardized rates are identical:

```{r same-risk}
same_risk <- tibble(
  age_group = c("0-19", "20-44", "45-64", "65+"),
  rate_per_person = c(0.0002, 0.0005, 0.0020, 0.0060),
  area_a = c(30000, 40000, 20000, 10000),
  area_b = c(15000, 26000, 30000, 30000)
) |>
  pivot_longer(c(area_a, area_b), names_to = "area", values_to = "population") |>
  mutate(cases = round(population * rate_per_person))

area_rates <- function(this_area) {
  area <- same_risk |> filter(area == this_area)
  islh_dsr_joined(
    area |> select(age_group, cases),
    area |> select(age_group, population),
    standard |> mutate(age_group = as.character(age_group)),
    by = "age_group",
    standard_id = standard$standard_id[1]
  ) |>
    transmute(crude = cases / population * 1e5, standardized = rate)
}

tibble(area = c("area_a", "area_b")) |>
  reframe(area_rates(area), .by = area) |>
  mutate(across(c(crude, standardized), \(x) round(x, 1)))
```

A crude rate is still the right number for planning services, because it
describes the burden in the population as it is. Report both, and say which
one a comparison uses.

## Suppress small cells

There is deliberately no default threshold. Supply the approved rule for the
specific release.

```{r suppress-vector}
counts <- c(0, 3, 5, 17, 42)
islh_suppress(counts, threshold = 5, label = "Suppressed")
```

Zero is kept because it is an observed absence, not a small positive count.
With the default `inclusive = TRUE`, both 3 and 5 are suppressed.

For a table, count the simulated cases and admissions in each North Island
local health area. The small areas have few cases, and fewer admissions:

```{r suppress-table}
area_counts <- islh_outbreak |>
  filter(hsda == "North Vancouver Island") |>
  summarise(
    cases = n(),
    admitted = sum(!is.na(date_admission)),
    .by = lha_name
  ) |>
  rename(area = lha_name) |>
  arrange(area)
area_counts

islh_suppress_table(
  area_counts,
  cols = c("cases", "admitted"),
  threshold = 5,
  label = "Suppressed"
)
```

A rate or confidence limit calculated from a hidden count gives the count
away: multiply the rate by the population and it is back. `linked` hides
those columns in the same rows:

```{r linked-suppression}
area_counts |>
  mutate(rate = round(admitted / cases * 100, 1)) |>
  islh_suppress_table(
    cols = "admitted",
    threshold = 5,
    linked = list(admitted = "rate")
  )
```

Linked cells become `NA` and keep the column numeric. Set `linked_label` to
show a neutral label instead. Either way, the audit records them with the
reason `"linked"`.

Complementary suppression can hide a second cell when one hidden value could
be recovered from a visible total:

```{r complementary-suppression}
islh_suppress_table(
  area_counts,
  cols = "cases",
  threshold = 5,
  label = "Suppressed",
  complementary = TRUE
)
```

By default the complementary rule looks at the whole column, which is right
for a table carrying one overall total. Where a table is subtotalled, name the
subtotal columns with `by` so the rule runs inside each block:

```{r grouped-suppression}
by_authority <- tibble(
  authority = c("Island", "Island", "Interior", "Interior"),
  area = c("North", "South", "East", "West"),
  cases = c(2, 30, 4, 25)
)

hidden <- islh_suppress_table(
  by_authority,
  cols = "cases",
  threshold = 5,
  label = "Suppressed",
  complementary = TRUE,
  by = "authority"
)

hidden
```

Column-wide, two cells are small and the arithmetic already fails to close.
Within each authority there is exactly one hidden cell, which a reader recovers
by subtracting from that authority's subtotal, so a second cell goes in each.

`islh_suppression_audit()` reports what was hidden and why, so a reviewer can
confirm the rule fired where it should have:

```{r suppression-audit}
islh_suppression_audit(hidden)
```

Keep the audit internal. It records no counts, but the reasons still say
something about the hidden values: a cell marked `"small"` is known to be at or
below the threshold. Publishing the audit, or a table that still carries it,
narrows what a reader can infer.

### Share a copy, not the working object

A result from this package is a working object. Besides the columns you see, it
can carry the suppression audit, the stratum counts behind a standardized rate
and the calendar settings of a count table. A printed table or a CSV of chosen
columns leaves those out, but an R object saved with `saveRDS()` keeps them.

`islh_suppress_table()` already removes stratum detail, because stratum counts
can reveal a hidden total. The audit stays, for your reviewers.
`islh_release_copy()` makes the copy to share: only the columns you name, and
none of the package's working information.

```{r release-copy}
audit <- islh_suppression_audit(hidden)
released <- islh_release_copy(hidden, c("authority", "area", "cases"))
released
names(attributes(released))
```

Keep `hidden` and `audit` with your working notes, and share `released`.

This helper works independently within each selected column. It does not prove
that a multidimensional or linked release is protected. Use `linked` for
rates, percentages and confidence limits calculated from a suppressed count.

Keep suppression separate from missing data and from precision. Without a
`label`, a suppressed cell and a missing value are both `NA`, so mark
suppressed cells in the table notes using the audit. A wide confidence
interval is a separate flag again: a count can be large enough to publish and
still too imprecise to compare.

## Round counts

```{r deterministic-rounding}
islh_round_base(c(0, 2, 3, 7, 12, 43), base = 5)
```

Deterministic rounding returns the same values each time. Random rounding is
unbiased over repeated use but varies between runs. Set a seed, round once and
save the released values:

```{r random-rounding}
set.seed(42)
islh_round_base(c(2, 3, 7, 12), base = 5, method = "random")
```

## Compare with an existing report

Before a new R output replaces a legacy report or a dashboard, check that the
two agree. `islh_compare_outputs()` lines the tables up on their keys and
returns one row for every value that differs, plus rows only one table has.
No rows means full agreement.

```{r compare-outputs}
r_output <- crude |>
  select(hsda, cases, rate)

# The dashboard rounds its rates, and counted one case differently.
dashboard <- r_output |>
  mutate(
    rate = round(rate, 1),
    cases = replace(cases, 1, cases[1] + 1)
  )

islh_compare_outputs(r_output, dashboard, by = "hsda", tolerance = 0.05)
```

`tolerance` absorbs rounding in the reference table's rates. It applies to
every numeric column, so keep it below 1 when counts are compared too, or
compare counts and rates in separate calls.

## Population and geography data

The BC Data Catalogue helpers require a network request and are documented in
[BC population and geography data](bcdata.html). That article shows the exact
returned columns, representative output, a checked join and troubleshooting
for unmatched rows.

Use `islandbrand` after the calculation when the result needs an Island Health
figure, table or Quarto report.

## References

Brown LD, Cai TT, DasGupta A (2001). Interval estimation for a binomial
proportion. *Statistical Science* 16(2):101-133.
<https://doi.org/10.1214/ss/1009213286>

Fay MP, Feuer EJ (1997). Confidence intervals for directly standardized rates:
a method based on the gamma distribution. *Statistics in Medicine*
16(7):791-801.
