---
title: "Routine surveillance with islandepi"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Routine surveillance with islandepi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

This guide demonstrates a small, complete workflow for daily or weekly
surveillance. Every chunk is evaluated when the pkgdown site is built, so the
tables below are actual package output. Where a function reports something
worth knowing, such as records it left out or periods it could not count, the
message is shown.

The package handles reusable mechanics. It does not decide case definitions,
alert policy or the public-health response to a statistical exceedance.

```{r packages}
library(dplyr)
library(tidyr)
library(islandepi)
```

The steps follow the order a report should:

1. Fix the report's cutoffs and the groups that should report.
2. Take the line list as it stood when it was extracted.
3. Check the records, the reporting delays and which feeds arrived.
4. Count events over the confirmed window.
5. Screen the most recent week that is nearly complete against a matched
   historical reference, and show the latest days as provisional.
6. Compare with earlier periods, with intervals.
7. Apply the release policy before anything leaves the team.

## Fix the cutoffs and the roster

Write down two dates before looking at the data:

- `report_end`, the last onset date the report covers;
- `extract_date`, the day the line list was extracted. Only cases reported by
  then are known.

They are often different: a report on onsets to Sunday may use an extract
taken on Tuesday. Here both are Sunday 1 February 2026.

Write down the groups that should appear, too. Deriving the cutoff from the
latest event date moves the report window backwards whenever the most recent
feeds are missing, and deriving the groups from the data drops any group that
sent nothing.

```{r cutoff}
report_end <- as.Date("2026-02-01")
extract_date <- as.Date("2026-02-01")

# Counting starts on Monday 3 November 2025, the first week with cases.
window_start <- as.Date("2025-11-03")

hsdas <- c(
  "South Vancouver Island",
  "Central Vancouver Island",
  "North Vancouver Island"
)
```

## Take the line list as it stood

`islh_outbreak` is a simulated line list bundled with the package: 790 cases
of a winter respiratory virus across Island Health, from November 2025 to
March 2026. It was simulated with the
[simulist](https://epiverse-trace.github.io/simulist/) package, so onsets,
reporting delays, ages and admissions behave like real surveillance data. No
record describes a real person.

It is the *final* line list, as it stood once every case had been reported.
A report written on 1 February could only see the cases reported by then.
Counting by onset date does not remove the rest: a case with onset on
30 January reported on 3 February has an onset inside the report window, but
nobody knew about it on 1 February. So restrict the line list on the report
date first.

In this example the North Island's files for 29 and 30 January also never
arrived, as the feed check below shows, so the cases those files carried are
missing from the extract too:

```{r event-data}
missing_files <- tibble(
  hsda = "North Vancouver Island",
  date_reported = as.Date(c("2026-01-29", "2026-01-30"))
)

events <- islh_outbreak |>
  filter(date_reported <= extract_date) |>
  anti_join(missing_files, by = join_by(hsda, date_reported))

events |>
  select(case_id, case_type, age, date_onset, date_reported, hsda) |>
  slice_head(n = 6)
```

For onsets in the report's last week, the difference is large:

```{r final-vs-known}
last_week <- bind_rows(
  final = islh_outbreak,
  known = events,
  .id = "line_list"
) |>
  filter(between(date_onset, report_end - 6, report_end)) |>
  count(hsda, line_list) |>
  pivot_wider(names_from = line_list, values_from = n)

last_week
```

The final line list holds `r sum(last_week$final)` cases with onset in that
week, but the extract on 1 February held only `r sum(last_week$known)`.
Counting from the final list would put cases in the report that nobody knew
about when it was written, and would make the latest week look further along
than it was.

## Check the records first

`islh_check_events()` returns one row for every problem. It does not alter the
line list or remove affected records.

```{r clean-event-check}
islh_check_events(
  events,
  id = case_id,
  date = date_onset,
  required = c(hsda, case_type),
  max_date = report_end
)
```

A deliberately damaged copy shows the review output:

```{r issue-example}
events_with_issues <- events |>
  slice_head(n = 5) |>
  mutate(
    case_id = replace(case_id, 2, case_id[1]),
    date_onset = replace(date_onset, 3, NA),
    hsda = replace(hsda, 4, NA)
  )

islh_check_events(
  events_with_issues,
  id = case_id,
  date = date_onset,
  required = hsda,
  max_date = report_end
)
```

Use the `.row` values to display the affected source records to the data owner.

Two more checks catch coding and entry errors. `allowed` lists the permitted
values of a coded field, and `date_order` names dates that must not come
before another, written `c(later = "earlier")`:

```{r issue-codes}
events_with_issues <- events_with_issues |>
  mutate(
    sex = replace(sex, 5, "F"),
    date_reported = replace(date_reported, 1, date_onset[1] - 2)
  )

islh_check_events(
  events_with_issues,
  id = case_id,
  date = date_onset,
  max_date = report_end,
  allowed = list(sex = c("Female", "Male")),
  date_order = c(date_reported = "date_onset")
)
```

When one person can have several legitimate records, such as encounters on
different days, pass the columns that together identify a record, for example
`id = c(person_id, encounter_date)`.

## Recent counts are incomplete

Counts by onset date fall away in the last few days because those cases have
not all been reported yet. `islh_reporting_delay()` summarizes the delays.
With `as_of`, it sees the data as they stood on that date:

```{r reporting-delay, message=TRUE}
islh_reporting_delay(
  events,
  onset = date_onset,
  report = date_reported,
  by = hsda,
  probs = c(0.5, 0.9, 0.99),
  as_of = extract_date,
  max_delay = 21
)
```

`max_delay` uses only onsets at least that many days old. Without it, the
recent cases, which are the quickly reported ones, make delays look shorter
than they are. Even with it, a case that takes longer than the time since its
onset has not arrived yet, so the summary describes the delays of cases
reported by `as_of`.

`islh_reporting_completeness()` turns delays into an expected share reported
for each recent onset date. It learns only from onsets at least `maturity`
days old, which you judge to be fully reported. That judgement comes from
knowing the reporting system, not from the data: a case that has not arrived
leaves no trace. A warning appears if older onsets in the data contradict it.

```{r reporting-completeness, message=TRUE}
completeness <- islh_reporting_completeness(
  events,
  onset = date_onset,
  report = date_reported,
  as_of = extract_date,
  max_delay = 14,
  maturity = 21
)

completeness |>
  select(onset_date, days_since_onset, reported, expected_complete) |>
  mutate(expected_complete = round(expected_complete, 2))
```

```{r completeness-facts, include=FALSE}
recent <- completeness |> filter(days_since_onset <= 3)
screen_end <- report_end - 7
screen_week <- screen_end - 6
screened <- completeness |>
  filter(between(onset_date, screen_week, screen_end))
```

The profile is learned from the `r completeness$delay_records[1]` cases with
onset between `r format(completeness$learn_from[1], "%e %B")` and
`r format(completeness$learn_to[1], "%e %B %Y")`. It rests on one more
assumption: **recent cases report like the cases it was learned from.** A
holiday, a surge that slows case entry, a new data feed or a change in
testing breaks that. The holidays sit inside this learning window, so if
reporting was slower then, the profile expects too little of January. Areas
can also differ: give `by = hsda` for a profile per area, which is unstable
for small areas (`delay_records` says how many cases each rests on), or
`pool = TRUE` to learn one profile for all of them. `learn_days` limits
learning to the most recent mature onsets, which follows a changing system
more closely but uses fewer cases.

`expected_complete` is an estimate from those cases, with uncertainty of its
own, not a known fraction. Use it to label the last days of an epidemic curve
as incomplete rather than read the drop as a decline. It is not a forecast of
the final count.

The last `r nrow(recent)` days are no more than
`r round(100 * max(recent$expected_complete))`% reported, and 1 February
barely at all. A week ending on the report date would be compared with
complete historical weeks while its own count is still rising, which biases
the screen toward "nothing unusual". So the screen below assesses the most
recent week that is nearly complete, `r format(screen_week, "%e")` to
`r format(screen_end, "%e %B")`, at least
`r floor(100 * min(screened$expected_complete))`% reported on every day, and
shows the latest days as provisional. When the latest days must be assessed,
use a method that adjusts for reporting delay, such as
`surveillance::bodaDelay()` (Salmon et al. 2015).

## Check which feeds arrived

A filled count table cannot tell a site that reported no cases from a site
whose data did not arrive. Both show zero. `islh_check_coverage()` compares a
log of received submissions, one row per file, extract or nil report, with the
groups that should report.

The log below is simulated: every HSDA sends one file a day, except that the
North Island's files for 29 and 30 January did not arrive.

```{r coverage, message=TRUE}
submissions <- expand_grid(
  hsda = hsdas,
  date = seq(window_start, extract_date, by = "day")
) |>
  anti_join(missing_files, by = join_by(hsda, date == date_reported))

feed_coverage <- islh_check_coverage(
  submissions,
  date = date,
  by = hsda,
  expected = hsdas,
  from = window_start,
  to = extract_date
)

feed_coverage |>
  filter(!received) |>
  select(hsda, period_start, received)
```

The files are dated by report, but the counts are by onset. A missing file
removes the cases reported that day, and their onsets fall over the days
before it. So an onset day's count is complete only when every file that
could carry its cases arrived: here, the files from the onset day to 14 days
later, since almost every case is reported within 14 days of onset (the 99th
percentile above). An inequality join finds those files for every onset day:

```{r onset-coverage, message=TRUE}
reach <- 14

onset_log <- expand_grid(
  hsda = hsdas,
  onset = seq(window_start, report_end, by = "day")
) |>
  mutate(last_file = pmin(onset + reach, extract_date)) |>
  inner_join(
    feed_coverage |> select(hsda, file_date = period_start, received),
    by = join_by(hsda, onset <= file_date, last_file >= file_date)
  ) |>
  summarise(received = all(received), .by = c(hsda, onset)) |>
  filter(received)

coverage <- islh_check_coverage(
  onset_log,
  date = onset,
  by = hsda,
  expected = hsdas,
  from = window_start,
  to = report_end
)

coverage_gaps <- coverage |>
  filter(!received) |>
  summarise(
    first = min(period_start),
    last = max(period_start),
    days = n(),
    .by = hsda
  )
coverage_gaps
```

Each row of `onset_log` stands for an onset day whose files all arrived, so
here the message counts onset days. Two missing days of files leave
`r coverage_gaps$days` days of the North Island's onset counts,
`r format(coverage_gaps$first, "%e")` to
`r format(coverage_gaps$last, "%e %B")`, incomplete. Treating only the onset
days 29 and 30 January as unknown would be simpler, and a reasonable rule of
thumb when delays are short, but it would call the North's screened week
complete when some of its cases were in the missing files.

Keep `coverage` with the report. The snapshot below uses it to show the North
Island's total as unknown. `complete` there means every expected file arrived
for those days. It does not mean their counts have stopped growing: that is
what the completeness profile above describes.

## Create complete daily counts

`groups` lists the areas that must appear even when an area has no cases on
some days. `fill = TRUE` also creates the days when nobody fell ill, so a quiet
day shows as a zero rather than a missing row. The window runs from Monday
3 November 2025 to Sunday 1 February 2026, whole weeks, so no events fall in
partial weeks later.

```{r daily-counts}
daily <- events |>
  islh_count_events(
    date = date_onset,
    id = case_id,
    by = hsda,
    interval = "day",
    from = window_start,
    to = report_end,
    fill = TRUE,
    groups = hsdas
  )

daily |> slice_head(n = 6)
```

For CDC epidemiological weeks, use `interval = "epiweek"`. ISO weeks use
`"isoweek"`. For a locally anchored week, use `"week"` and provide
`week_start`, where 1 is Monday and 7 is Sunday.

```{r weekly-counts}
weekly <- daily |>
  islh_count_events(
    date = period_start,
    count = count,
    by = hsda,
    interval = "week",
    week_start = 1,
    from = window_start,
    to = report_end,
    fill = TRUE
  )

weekly |> slice_head(n = 6)
```

The result uses period start and end dates rather than a hand-built week number.
That removes special January and December corrections from report scripts.

Summing daily counts into weeks is safe here because each case has one record.
With `id`, `islh_count_events()` counts distinct people within each day, so a
person with records on two days of the same week would be counted twice in
the weekly sum. The daily result records how many identifiers appear on more
than one day, and summing it warns when that number is not zero. To count
people per week in that case, recount the line list with
`interval = "week"`.

## Calculate a transparent baseline

A reference period should describe what is usual for the time of year. For a
seasonal illness, that means the same weeks in earlier seasons, often with a
few weeks either side. `islh_reference_periods()` lists them for the week
being screened.

`islh_seasons` holds seven seasons of weekly counts. Its 2025-26 season is
this outbreak, and the earlier seasons are simulated on the same scale. The
2020-21 season is left out here. Public health measures that winter changed
contact patterns, testing and reporting, so its weeks do not describe a usual
season. Record a decision like that in the report's methods, and base it on
what changed in the population or the reporting system, not on the counts
alone: leaving out a season because it looks low, or high, tunes the
reference to the answer.

```{r matched-reference}
matched_weeks <- islh_reference_periods(
  screen_week,
  years_back = 5,
  window = 2,
  exclude = data.frame(from = "2020-09-01", to = "2021-08-31")
)
range(matched_weeks)
length(matched_weeks)

matched <- islh_surveillance_baseline(
  islh_seasons,
  date = period_start,
  value = count,
  by = hsda,
  reference_periods = matched_weeks
)

matched |>
  select(hsda, reference_n, reference_mean, reference_sd, upper_limit)
```

ISO and epidemiological weeks keep their week number from year to year, and
ordinary weeks take the nearest week a year back, so the reference does not
drift by a day each year.

The matched reference describes the same weeks of earlier seasons, so it
answers the question a winter report asks: is this late January busier than
late January usually is? It is also wide:

```{r reference-spread}
reference_spread <- islh_seasons |>
  filter(period_start %in% matched_weeks) |>
  summarise(
    weeks = n(),
    lowest = min(count),
    highest = max(count),
    mean = round(mean(count), 1),
    .by = hsda
  )
reference_spread
```

```{r reference-facts, include=FALSE}
south_spread <- reference_spread |> filter(hsda == "South Vancouver Island")
south_matched <- matched |> filter(hsda == "South Vancouver Island")
```

The South Island's reference weeks run from `r south_spread$lowest` to
`r south_spread$highest` cases, because each season peaks at a different
height and in a different week. The default screen, the mean plus two
standard deviations, gives a limit of `r round(south_matched$upper_limit)`,
`r format(round(south_matched$upper_limit / south_matched$reference_mean, 1), nsmall = 1)`
times the mean, so a week has to be far above usual before it is flagged.
That screen has low sensitivity on skewed, seasonal counts.
`method = "quantile"` describes the spread without assuming it is
symmetric, and the model-based method below allows for trend and
seasonality.

### A counterexample: an early-season reference

A reference from earlier in the same season answers a different question. The
next chunk uses the first eight weeks of this season, November and December,
only to show the difference. Do not copy it for a seasonal disease: it
compares peak-season weeks with the start of the season, so it flags weeks
that are ordinary for the time of year.

```{r baseline}
early_weeks <- weekly |>
  distinct(period_start) |>
  slice_min(period_start, n = 8) |>
  pull(period_start)

baseline <- islh_surveillance_baseline(
  weekly,
  date = period_start,
  value = count,
  by = hsda,
  reference_periods = early_weeks,
  method = "mean_sd",
  multiplier = 2,
  prediction_adjustment = TRUE,
  minimum_periods = 4
)

baseline |>
  select(hsda, reference_n, reference_mean, upper_limit)
```

`reference_n` and `reference_adjustment` make the calculation auditable. The
adjustment is derived from the number of periods rather than hardcoded.
`method = "range"` returns the historical minimum and maximum, while
`method = "quantile"` uses the supplied probabilities.

The default screen is strictly above `upper_limit`. Choose
`comparison = "at_or_above"` explicitly to include equality. A zero current
count against a zero limit is unflagged by default. These are descriptive
screens, not calibrated outbreak probabilities or 95% prediction limits.
Mean-SD references require at least two observations, and all requested
reference periods must be present and complete. Reference dates and method
parameters are retained in `attr(baseline, "islh_reference")`.

## Screen the most recent complete week

The snapshot puts the seven days of the screened week beside the matched
reference, with the coverage check deciding which counts are known:

```{r snapshot, message=TRUE}
snapshot <- islh_surveillance_snapshot(
  daily,
  date = period_start,
  value = count,
  by = hsda,
  end = screen_end,
  periods = 7,
  baseline = matched,
  coverage = coverage,
  include_total = TRUE
)

snapshot |>
  select(hsda, total, complete, upper_limit, exceeds_reference)
```

```{r snapshot-facts, include=FALSE}
known <- snapshot |> filter(hsda %in% hsdas, complete)
flagged <- known |> filter(exceeds_reference %in% TRUE) |> pull(hsda)
```

The North Island's total is unknown, because files that could carry its
cases did not arrive, and so is the `All` row: a total that silently missed
those cases would understate the week. Of the two areas with complete data,
`r if (length(flagged) == 0) "neither is" else paste(flagged, collapse = " and ")`
`r if (length(flagged) == 1) "is" else if (length(flagged) > 1) "are"` above
the upper limit for this time of year in earlier seasons. A flag is a signal
to look at, not a finding.

Against the early-season reference, the same week looks different:

```{r snapshot-early}
early <- islh_surveillance_snapshot(
  daily,
  date = period_start,
  value = count,
  by = hsda,
  end = screen_end,
  periods = 7,
  baseline = baseline,
  coverage = coverage
)

early |>
  select(hsda, total, early_limit = upper_limit, early_flag = exceeds_reference) |>
  left_join(
    matched |> select(hsda, matched_limit = upper_limit),
    by = join_by(hsda)
  ) |>
  mutate(across(ends_with("limit"), \(x) round(x, 1)))
```

```{r early-facts, include=FALSE}
south_early <- early |> filter(hsda == "South Vancouver Island")
```

The early-season limits are a fraction of the matched ones, because November
is quieter than late January. The South Island's `r south_early$total` cases
sit just under its early-season limit of
`r round(south_early$upper_limit, 1)`, and far under the matched limit of
`r round(south_matched$upper_limit)`. Against a reference from the wrong time
of year, an ordinary late-January week looks borderline, and a slightly
busier one would be flagged. The full table also holds the seven daily
columns and every reference statistic.

The seven displayed daily counts add to one weekly total, so the weekly
baseline above is comparable with `total`. A seven-day total must not be
compared with a baseline calculated from individual daily counts, and the
function refuses to: both tables carry the reporting period they were built
with, and mismatched durations stop with an error rather than producing a table
whose limits are seven times too small. If you display a different window,
calculate the baseline from historical totals of that same duration.

Metadata travels as attributes, and transformations may keep stale attributes
or drop them. The functions check the actual period alignment and coverage. If
you reshape a table between steps, name the reporting period explicitly: pass
`interval` to `islh_surveillance_baseline()`, or `baseline_interval` to
`islh_surveillance_snapshot()` for limits that came from somewhere else
entirely, such as last season's totals held in a spreadsheet.

The daily columns, `total` and baseline fields are ordinary data. Export them
with the package required by the report, or pass the result to
`islandbrand::islh_gt()` or `islandbrand::islh_flextable()`.

Only use `include_total = TRUE` when the groups are mutually exclusive. If a
record can contribute to more than one group, calculate the overall total from
the unduplicated event data instead.

### The latest days, as provisional

Show the days after the screened week with how complete they are expected to
be, and without a flag:

```{r provisional}
daily |>
  filter(period_start > screen_end) |>
  left_join(
    coverage |> select(hsda, period_start, received),
    by = join_by(hsda, period_start)
  ) |>
  mutate(
    count = if_else(received, count, NA),
    hsda = sub(" Vancouver Island", "", hsda)
  ) |>
  select(onset_date = period_start, hsda, count) |>
  pivot_wider(names_from = hsda, values_from = count) |>
  left_join(
    completeness |> select(onset_date, expected_complete),
    by = join_by(onset_date)
  ) |>
  mutate(expected_complete = round(expected_complete, 2))
```

## Compare with earlier periods

`islh_compare_periods()` puts a week beside the previous week, the same week
last year, or the season so far against the same stretch of last season.
Every row names the dates it covers.

`islh_seasons` holds the final counts of every season, this one included. For
a report written on 1 February, take the earlier seasons from it and this
season from the cases reported by then. `bind_rows()` keeps the first table's
metadata, which no longer describes the joined table, so the calls that
follow name the interval:

```{r series}
current_season <- events |>
  islh_count_events(
    date = date_onset,
    id = case_id,
    by = hsda,
    interval = "week",
    from = "2025-09-01",
    to = report_end,
    fill = TRUE,
    groups = hsdas
  )

series <- islh_seasons |>
  filter(period_start < as.Date("2025-09-01")) |>
  bind_rows(current_season)
```

```{r compare-periods}
comparison <- islh_compare_periods(
  series,
  date = period_start,
  value = count,
  by = hsda,
  current = screen_week,
  comparison = c("previous_period", "previous_year", "season_to_date"),
  season_start = "2025-09-01",
  interval = "week",
  week_start = 1
)

comparison |>
  as_tibble() |>
  filter(hsda != "North Vancouver Island") |>
  select(hsda, comparison, current, reference, ratio, ratio_lower, ratio_upper) |>
  mutate(across(starts_with("ratio"), \(x) round(x, 2)))
```

```{r compare-facts, include=FALSE}
south_year <- comparison |>
  filter(hsda == "South Vancouver Island", comparison == "previous_year")
```

The North Island is left out: its counts for the week are incomplete, because
of the missing files.

`ratio_lower` and `ratio_upper` are exact 95% limits for the ratio under a
Poisson model, as `islh_rate_ratio()` gives them. The South Island had
`r south_year$current` cases against `r south_year$reference` in the same
week last year, a ratio of `r round(south_year$ratio, 2)`. That is an
increase, estimated imprecisely: the counts are compatible with anything from
almost no change to about `r round(south_year$ratio_upper, 1)` times last
year's level. Read the limits as the least uncertainty there is. Cases in an
outbreak cluster, so they vary more than Poisson counts, and a change in
testing or case definition between the two years is not in the interval at
all.

With small counts, lead with the counts: "9 cases against 2" says more than a
350% rise. `ratio` and `percent_change` are `NA` when the reference is zero,
because a rise from 0 to 3 cases has no size as a percentage. A missing week
in either window gives a missing comparison rather than a smaller total.

`islh_reporting_calendar()` gives any date its week-year, fiscal-year and
season labels, using the same rules:

```{r reporting-calendar}
islh_reporting_calendar(
  c("2025-12-28", "2025-12-29", "2026-01-26"),
  season_start_month = 9
)
```

## Model-based detection

The baselines above are descriptive screens. `islh_farrington()` runs the
improved Farrington algorithm from the surveillance package on weekly counts:
for each week it fits a quasi-Poisson model to comparable weeks in earlier
years, allowing for trend, seasonality and overdispersion, and flags a count
above the model's upper limit.

`alpha` sets that limit, and it is nominal. The default threshold,
`"nbPlugin"`, plugs the estimated negative binomial parameters into the
limit without allowing for the uncertainty in estimating them, so
`alpha = 0.05` is not the false-alarm rate you will see in practice. The
method also needs several years of history.

The same comparability decision applies here. With `years_back = 5`, weeks
in January 2026 would learn from the 2020-21 season, which the matched
reference above leaves out. `years_back = 4` carries that decision forward:

```{r farrington, eval=requireNamespace("surveillance", quietly = TRUE)}
detection <- islh_farrington(
  series,
  date = period_start,
  value = count,
  by = hsda,
  from = "2025-12-29",
  to = screen_week,
  years_back = 4
)

detection |>
  as_tibble() |>
  filter(hsda == "South Vancouver Island") |>
  select(period_start, observed, expected, upper_limit, alarm, status) |>
  mutate(expected = round(expected, 1))
```

```{r farrington-facts, include=FALSE, eval=requireNamespace("surveillance", quietly = TRUE)}
with_2020 <- islh_farrington(
  series,
  date = period_start,
  value = count,
  by = hsda,
  from = screen_week,
  years_back = 5
) |>
  filter(hsda == "South Vancouver Island")
without_2020 <- detection |>
  filter(hsda == "South Vancouver Island", period_start == screen_week)
```

```{r farrington-text, echo=FALSE, results="asis", eval=requireNamespace("surveillance", quietly = TRUE)}
cat(
  "For the week of ", format(screen_week, "%e %B"), ", the South Island's ",
  "expected count is ", round(without_2020$expected, 1), " and its limit ",
  round(without_2020$upper_limit), ". With 2020-21 in the history they would ",
  "be ", round(with_2020$expected, 1), " and ", round(with_2020$upper_limit),
  ": the quiet season pulls the expected count down. Decide which seasons ",
  "are comparable once, and apply the decision to every method.",
  sep = ""
)
```

`status` says whether each week was assessed. A week with fewer than 5 cases in
the last 4 weeks is set aside by the method's low-count rule, and shows
`"low_count_rule"` with a missing `alarm`: it was not assessed, which is not the
same as quiet. Show the status in any table of results. `low_count` and
`low_count_weeks` change the rule, and the settings used are recorded in
`attr(detection, "islh_farrington")`.

Assess only weeks that are nearly complete, as here. The model compares a
week with complete earlier weeks, so a week still being reported looks quiet.

Evaluate the method locally before relying on it. Run it over past seasons
and record, for events the team knows about:

- **alert burden:** how many alarms it raises;
- **sensitivity:** how many of the events it flags;
- **timeliness:** how many weeks after an event starts the first alarm comes;
- **positive predictive value:** how many alarms were real events.

Then set `alpha`. Tuning it to reduce alarms alone can hide a method that
misses events or finds them late.

Partial weeks are refused, in the weeks assessed and in the history alike,
because a partial week's count does not describe the week its date names.
`islh_compare_periods()` refuses them for the same reason.

Install surveillance with `install.packages("surveillance", type = "binary")`
to use it. The rest of `islandepi` does not need it.

## Encounters or people?

`islh_encounters` holds four weeks of simulated encounters from three sites.
Some people come back, and three records were resent in a later file. First
find the resent records:

```{r encounters-check}
issues <- islh_check_events(
  islh_encounters,
  id = encounter_id,
  date = encounter_date,
  max_date = "2026-08-30"
)
issues |> count(.issue)

unique_encounters <- islh_encounters |> distinct()
```

Counting rows counts encounters; counting `id` counts people. Distinct people
per day cannot simply be added up into distinct people per week, because a
person seen on two days of a week would count twice. The daily count records
how many people appear on more than one day, and summing it warns:

```{r encounters-people, warning=TRUE}
daily_people <- unique_encounters |>
  islh_count_events(
    date = encounter_date,
    id = person_id,
    from = "2026-08-03",
    to = "2026-08-30"
  )

weekly_sum <- daily_people |>
  islh_count_events(
    date = period_start,
    count = count,
    interval = "week",
    from = "2026-08-03",
    to = "2026-08-30"
  )
```

To count people per week, count the line list at the weekly interval:

```{r encounters-weekly}
weekly_people <- unique_encounters |>
  islh_count_events(
    date = encounter_date,
    id = person_id,
    interval = "week",
    from = "2026-08-03",
    to = "2026-08-30"
  )

tibble(
  week = weekly_people$period_start,
  summed_daily = weekly_sum$count,
  people = weekly_people$count
)
```

`islh_feed_log` records which daily files arrived. Site C's files for 12 and
13 August are missing, while Site B sent nil reports on days it had no
encounters. Suppose a fourth site, Site D, joined the roster this month but
has not sent a single file. `islh_check_coverage()` tells all three situations
apart:

```{r encounters-coverage, message=TRUE}
sites <- c("Site A", "Site B", "Site C", "Site D")

site_coverage <- islh_check_coverage(
  islh_feed_log,
  date = date,
  by = site,
  expected = sites,
  from = "2026-08-03",
  to = "2026-08-30"
)

site_coverage |>
  filter(!received) |>
  count(site)
```

Site D has no encounters at all, so it has no rows in the counts. A snapshot
built from the counts alone would leave it out, and its `All` row would look
complete. With `coverage`, Site D stays in the table as unknown:

```{r encounters-snapshot, message=TRUE}
site_counts <- unique_encounters |>
  islh_count_events(
    date = encounter_date,
    by = site,
    from = "2026-08-03",
    to = "2026-08-30",
    groups = sites
  )

site_week <- islh_surveillance_snapshot(
  site_counts,
  date = period_start,
  value = count,
  by = site,
  end = "2026-08-16",
  periods = 7,
  include_total = TRUE,
  coverage = site_coverage
)

site_week |> select(site, total, complete)
```

Site C is unknown for the week because two files are missing. Site D is
unknown because nothing arrived, even though `groups` filled its days with
zeros. The `All` total is unknown too.

## Plot the counts

`islandepi` intentionally returns data rather than imposing a figure style. Use
`islandbrand::islh_epi_curve()` to create the Island Health figure:

```{r plot-setup, include=FALSE, eval=requireNamespace("islandbrand", quietly = TRUE)}
# R CMD check replays this code on the pdf device, which stops on a font
# family it does not know. islh_setup() picks a family that is installed.
islandbrand::islh_setup(format = "plots", quiet = TRUE)
```

```{r plot-example, eval=requireNamespace("islandbrand", quietly = TRUE), fig.width=8, fig.height=4.5, fig.alt="Stacked weekly case counts by health service delivery area from 3 November 2025 to 1 February 2026, counting the cases reported by 1 February. Cases rise to a first peak of 72 in the week of 8 December, dip over the holidays and peak again at 75 in the week of 19 January. The last week shows 43 cases and is incomplete. South Vancouver Island has the most cases each week."}
weekly |>
  # Short labels keep the legend on one line.
  mutate(area = sub(" Vancouver Island", " Island", hsda)) |>
  islandbrand::islh_epi_curve(
    date = period_start,
    count = count,
    fill = area,
    labels = "total",
    title = "Weekly cases by health service delivery area",
    subtitle = "By week of symptom onset, cases reported by 1 February 2026",
    caption = paste(
      "The last week is incomplete.",
      "Source: simulated outbreak (islh_outbreak). Not real surveillance data."
    )
  ) +
  ggplot2::labs(fill = NULL)
```

The published package website installs `islandbrand` and evaluates this example.
When this vignette is built on a computer without `islandbrand`, only the plot
is skipped; all analytical examples still run.

## Coverage and reporting calendar

Timestamp conversion defaults to `America/Vancouver`; supply `timezone` to
change the reporting calendar explicitly. Both the first and last incomplete
periods are excluded by default from counts, and a message says how many
events that left out. Preaggregated weekly counts cannot be relabelled to a
different week start.

Snapshots fail on missing source periods by default. Use
`missing_periods = "missing"` to propagate unknown values and `complete = FALSE`.
Use `missing_periods = "zero"` only after confirming those absent records mean
zero activity; that decision is recorded in `islh_screening`. Event-grid
completion likewise assumes that `from` and `to` describe confirmed coverage;
use `islh_check_coverage()` to confirm it.

Grouping columns must not share a name with a column the functions create,
such as `total`, `complete` or `upper_limit`. Rename them first; the functions
stop rather than overwrite the group labels.

A table without the metadata `islh_count_events()` records must say what one
row covers: pass `source_interval` to the snapshot, or `interval` to the
baseline and comparison functions. The functions only work it out themselves
for consecutive days, because rows a week apart could be weekly totals or
daily counts with days missing.

## Before anything leaves the team

Apply the approved suppression rule with `islh_suppress_table()`, then share a
copy made with `islh_release_copy()`, which keeps only the columns you name
and none of the package's working information. Keep the working objects, the
coverage check and the suppression audit with the report's working notes.

## References

Charniga K, Park SW, Akhmetzhanov AR, et al. (2024). Best practices for
estimating and reporting epidemiological delay distributions of infectious
diseases. *PLOS Computational Biology* 20(10):e1012520.
<https://doi.org/10.1371/journal.pcbi.1012520>

Noufaily A, Enki DG, Farrington P, Garthwaite P, Andrews N, Charlett A (2013).
An improved algorithm for outbreak detection in multiple surveillance systems.
*Statistics in Medicine* 32(7):1206-1222.
<https://doi.org/10.1002/sim.5595>

Salmon M, Schumacher D, Stark K, Höhle M (2015). Bayesian outbreak detection
in the presence of reporting delays. *Biometrical Journal* 57(6):1051-1067.
<https://doi.org/10.1002/bimj.201400159>

Salmon M, Schumacher D, Höhle M (2016). Monitoring count time series in R:
aberration detection in public health surveillance. *Journal of Statistical
Software* 70(10):1-35. <https://doi.org/10.18637/jss.v070.i10>
