---
title: "Routine surveillance with islandepi"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Routine surveillance with islandepi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(islandepi)
```

This guide demonstrates a small, complete workflow for daily or weekly
surveillance. Every chunk is evaluated when the pkgdown site is built, so the
tables below are actual package output.

The package handles reusable mechanics. It does not decide case definitions,
alert policy or the public-health response to a statistical exceedance.

## Example event data

`islh_outbreak` is a simulated line list bundled with the package: 790 cases
of a winter respiratory virus across Island Health, from November 2025 to
March 2026. It was simulated with the
[simulist](https://epiverse-trace.github.io/simulist/) package, so onsets,
reporting delays, ages and admissions behave like real surveillance data. No
record describes a real person.

```{r event-data}
events <- islh_outbreak
head(events)
```

## Check the records first

`islh_check_events()` returns one row for every problem. It does not alter the
line list or remove affected records.

```{r clean-event-check}
islh_check_events(
  events,
  id = case_id,
  date = date_onset,
  required = c(hsda, case_type),
  max_date = "2026-03-08"
)
```

A deliberately damaged copy shows the review output:

```{r issue-example}
events_with_issues <- events[1:5, ]
events_with_issues$case_id[2] <- events_with_issues$case_id[1]
events_with_issues$date_onset[3] <- NA
events_with_issues$hsda[4] <- NA

islh_check_events(
  events_with_issues,
  id = case_id,
  date = date_onset,
  required = hsda,
  max_date = "2026-03-08"
)
```

Use the `.row` values to display the affected source records to the data owner.

Two more checks catch coding and entry errors. `allowed` lists the permitted
values of a coded field, and `date_order` names dates that must not come
before another, written `c(later = "earlier")`:

```{r issue-codes}
events_with_issues$sex[5] <- "F"
events_with_issues$date_reported[1] <- events_with_issues$date_onset[1] - 2

islh_check_events(
  events_with_issues,
  id = case_id,
  date = date_onset,
  max_date = "2026-03-08",
  allowed = list(sex = c("Female", "Male")),
  date_order = c(date_reported = "date_onset")
)
```

When one person can have several legitimate records, such as encounters on
different days, pass the columns that together identify a record, for example
`id = c(person_id, encounter_date)`.

## Create complete daily counts

`groups` lists the areas that must appear even when an area has no cases on
some days. `fill = TRUE` also creates the days when nobody fell ill, so a quiet
day shows as a zero rather than a missing row. The window runs from Monday
3 November 2025 to Sunday 8 March 2026, whole weeks, so no events fall in
partial weeks later.

```{r daily-counts}
hsdas <- c(
  "South Vancouver Island",
  "Central Vancouver Island",
  "North Vancouver Island"
)

daily <- islh_count_events(
  events,
  date = date_onset,
  id = case_id,
  by = hsda,
  interval = "day",
  from = "2025-11-03",
  to = "2026-03-08",
  fill = TRUE,
  groups = hsdas
)

head(daily, 6)
```

For CDC epidemiological weeks, use `interval = "epiweek"`. ISO weeks use
`"isoweek"`. For a locally anchored week, use `"week"` and provide
`week_start`, where 1 is Monday and 7 is Sunday.

```{r weekly-counts}
weekly <- islh_count_events(
  daily,
  date = period_start,
  count = count,
  by = hsda,
  interval = "week",
  week_start = 1,
  from = "2025-11-03",
  to = "2026-03-08",
  fill = TRUE
)

head(weekly, 6)
```

The result uses period start and end dates rather than a hand-built week number.
That removes special January and December corrections from report scripts.

Summing daily counts into weeks is safe here because each case has one record.
With `id`, `islh_count_events()` counts distinct people within each day, so a
person with records on two days of the same week would be counted twice in
the weekly sum. The daily result records how many identifiers appear on more
than one day, and summing it warns when that number is not zero. To count
people per week in that case, recount the line list with
`interval = "week"`.

## Calculate a transparent baseline

A reference period should describe what is usual for the time of year. For a
seasonal illness, that means the same weeks in earlier seasons, often with a
few weeks either side. `islh_reference_periods()` lists them. Leave out
seasons that were not usual, and say so in the report.

`islh_seasons` holds seven simulated seasons of weekly counts. Its 2020-21
season barely rose, so it is excluded here:

```{r matched-reference}
matched_weeks <- islh_reference_periods(
  "2026-01-26",
  years_back = 5,
  window = 2,
  exclude = data.frame(from = "2020-09-01", to = "2021-08-31")
)
range(matched_weeks)
length(matched_weeks)

matched <- islh_surveillance_baseline(
  islh_seasons,
  date = period_start,
  value = count,
  by = hsda,
  reference_periods = matched_weeks
)
matched[c("hsda", "reference_n", "reference_mean", "upper_limit")]
```

ISO and epidemiological weeks keep their week number from year to year, and
ordinary weeks take the nearest week a year back, so the reference does not
drift by a day each year.

`islh_outbreak` covers one season only, so the snapshot example below uses its
first eight weeks, November and December, as a stand-in reference. That
choice is for illustration: it compares peak-season weeks with the start of
the season, so it flags more weeks than a matched reference would.

```{r baseline}
reference_weeks <- sort(unique(weekly$period_start))[1:8]

baseline <- islh_surveillance_baseline(
  weekly,
  date = period_start,
  value = count,
  by = hsda,
  reference_periods = reference_weeks,
  method = "mean_sd",
  multiplier = 2,
  prediction_adjustment = TRUE,
  minimum_periods = 4
)

baseline[c("hsda", "reference_n", "reference_mean", "upper_limit")]
```

`reference_n` and `reference_adjustment` make the calculation auditable. The
adjustment is derived from the number of periods rather than hardcoded.
`method = "range"` returns the historical minimum and maximum, while
`method = "quantile"` uses the supplied probabilities.

The default screen is strictly above `upper_limit`. Choose
`comparison = "at_or_above"` explicitly to include equality. A zero current
count against a zero limit is unflagged by default. These are descriptive
screens, not calibrated outbreak probabilities or 95% prediction limits.
Mean-SD references require at least two observations, and all requested
reference periods must be present and complete. Reference dates and method
parameters are retained in `attr(baseline, "islh_reference")`.

## Build the current seven-day table

```{r snapshot}
snapshot <- islh_surveillance_snapshot(
  daily,
  date = period_start,
  value = count,
  by = hsda,
  end = "2026-02-01",
  periods = 7,
  baseline = baseline,
  include_total = FALSE
)

snapshot[c("hsda", "total", "upper_limit", "exceeds_reference")]
```

In the week of 26 January, near the peak of the wave, the North Island's
total is above its early-season upper limit, and the South Island is just
under its own. That is what the screen is for: a signal to look at, not a
finding. Against a matched reference from earlier seasons, a winter peak would
only stand out if it were larger than usual for late January. The full table also holds the seven daily columns and every
reference statistic.

The seven displayed daily counts add to one weekly total, so the weekly
baseline above is comparable with `total`. A seven-day total must not be
compared with a baseline calculated from individual daily counts, and the
function refuses to: both tables carry the reporting period they were built
with, and mismatched durations stop with an error rather than producing a table
whose limits are seven times too small. If you display a different window,
calculate the baseline from historical totals of that same duration.

Metadata travels as attributes; transformations may retain stale attributes
or discard them. Functions validate actual period alignment and coverage. If you reshape a table between steps, name the reporting period
explicitly: pass `interval` to `islh_surveillance_baseline()`, or
`baseline_interval` to `islh_surveillance_snapshot()` for limits that came from
somewhere else entirely, such as last season's totals held in a spreadsheet.

The daily columns, `total` and baseline fields are ordinary data. Export them
with the package required by the report, or pass the result to
`islandbrand::islh_gt()` or `islandbrand::islh_flextable()`.

Only use `include_total = TRUE` when the groups are mutually exclusive. If a
record can contribute to more than one group, calculate the overall total from
the unduplicated event data instead.

## Separate missing feeds from quiet days

A filled count table cannot tell a site that reported no cases from a site
whose data did not arrive. Both show zero. `islh_check_coverage()` compares a
log of received submissions, one row per file, extract or nil report, with the
sites that should report.

The log below is simulated: every HSDA sends one file a day, except that the
North Island's files for 29 and 30 January did not arrive.

```{r coverage}
days <- seq(as.Date("2026-01-26"), as.Date("2026-02-01"), by = "day")
submissions <- expand.grid(hsda = hsdas, date = days, stringsAsFactors = FALSE)
missing_files <- submissions$hsda == "North Vancouver Island" &
  submissions$date %in% as.Date(c("2026-01-29", "2026-01-30"))
submissions <- submissions[!missing_files, ]

coverage <- islh_check_coverage(
  submissions,
  date = date,
  by = hsda,
  expected = hsdas,
  from = "2026-01-26",
  to = "2026-02-01"
)
coverage[!coverage$received, c("hsda", "period_start", "received")]
```

Remove the periods that were not received from the counts, then build the
snapshot with `missing_periods = "missing"`. The North Island now shows an
unknown total rather than a total that quietly misses two days:

```{r coverage-snapshot}
received <- coverage[coverage$received, c("hsda", "period_start")]
known <- merge(daily, received)

gated <- islh_surveillance_snapshot(
  known,
  date = period_start,
  value = count,
  by = hsda,
  end = "2026-02-01",
  periods = 7,
  interval = "day",
  baseline = baseline,
  baseline_interval = "week",
  missing_periods = "missing"
)

gated[c("hsda", "total", "complete", "exceeds_reference")]
```

`merge()` drops the reporting-period metadata, so the snapshot takes the
interval from the dates and the baseline duration from `baseline_interval`.

## Compare with earlier periods

`islh_compare_periods()` puts the current week beside the previous week, the
same week last year, or the season so far against the same stretch of last
season. Every row names the dates it covers:

```{r compare-periods}
comparison <- islh_compare_periods(
  islh_seasons,
  date = period_start,
  value = count,
  by = hsda,
  current = "2026-01-26",
  comparison = c("previous_period", "previous_year", "season_to_date"),
  season_start = "2025-09-01"
)
comparison[
  comparison$hsda == "North Vancouver Island",
  c("comparison", "reference_from", "current", "reference", "percent_change")
]
```

`ratio` and `percent_change` are `NA` when the reference is zero: a rise from
0 to 3 cases has no size as a percentage, so report the counts. A missing week
in either window gives a missing comparison rather than a smaller total.

`islh_reporting_calendar()` gives any date its week-year, fiscal-year and
season labels, using the same rules:

```{r reporting-calendar}
islh_reporting_calendar(
  c("2025-12-28", "2025-12-29", "2026-01-26"),
  season_start_month = 9
)
```

## Recent counts are incomplete

Counts by onset date fall away in the last few days because those cases have
not all been reported yet. `islh_reporting_delay()` summarizes the delays.
With `as_of`, it sees the data as they stood on that date, leaving out cases
reported later:

```{r reporting-delay}
islh_reporting_delay(
  events,
  onset = date_onset,
  report = date_reported,
  by = hsda,
  as_of = "2026-01-20",
  max_delay = 10
)
```

`max_delay` uses only onsets old enough for every delay up to that many days
to have been seen. Without it, the recent cases, which are the quickly
reported ones, make delays look shorter than they are.

`islh_reporting_completeness()` turns those delays into an expected share
reported for each recent onset date:

```{r reporting-completeness}
completeness <- islh_reporting_completeness(
  events,
  onset = date_onset,
  report = date_reported,
  as_of = "2026-01-20",
  max_delay = 10
)
completeness$expected_complete <- round(completeness$expected_complete, 2)
tail(completeness, 5)
```

On 20 January, cases with onset on 18 January were only about half reported.
Label the last few days of an epidemic curve as incomplete rather than read
the drop as a decline.

## Encounters or people?

`islh_encounters` holds four weeks of simulated encounters from three sites.
Some people come back, and three records were resent in a later file. First
find the resent records:

```{r encounters-check}
issues <- islh_check_events(
  islh_encounters,
  id = encounter_id,
  date = encounter_date,
  max_date = "2026-08-30"
)
table(issues$.issue)
unique_encounters <- islh_encounters[!duplicated(islh_encounters), ]
```

Counting rows counts encounters; counting `id` counts people. Distinct people
per day cannot simply be added up into distinct people per week, because a
person seen on two days of a week would count twice. The daily count records
how many people appear on more than one day, and summing it warns:

```{r encounters-people, warning=TRUE}
daily_people <- islh_count_events(
  unique_encounters,
  date = encounter_date,
  id = person_id,
  from = "2026-08-03",
  to = "2026-08-30"
)

weekly_sum <- islh_count_events(
  daily_people,
  date = period_start,
  count = count,
  interval = "week",
  from = "2026-08-03",
  to = "2026-08-30"
)
```

To count people per week, count the line list at the weekly interval:

```{r encounters-weekly}
weekly_people <- islh_count_events(
  unique_encounters,
  date = encounter_date,
  id = person_id,
  interval = "week",
  from = "2026-08-03",
  to = "2026-08-30"
)
data.frame(
  week = weekly_people$period_start,
  summed_daily = weekly_sum$count,
  people = weekly_people$count
)
```

`islh_feed_log` records which daily files arrived. Site C's files for 12 and
13 August are missing, while Site B sent nil reports on days it had no
encounters. `islh_check_coverage()` tells them apart:

```{r encounters-coverage}
feed_coverage <- islh_check_coverage(
  islh_feed_log,
  date = date,
  by = site,
  expected = c("Site A", "Site B", "Site C"),
  from = "2026-08-03",
  to = "2026-08-30"
)
feed_coverage[!feed_coverage$received, c("site", "period_start")]
```

## Model-based detection

The baselines above are descriptive screens. For a detection method with a
stated false-alarm rate, `islh_farrington()` runs the improved Farrington
algorithm from the surveillance package on weekly counts. It needs several
years of history, and it needs local evaluation: run it over past seasons,
count the alarms, and set `alpha` so the alert burden is workable.

```{r farrington, eval=requireNamespace("surveillance", quietly = TRUE)}
detection <- islh_farrington(
  islh_seasons,
  date = period_start,
  value = count,
  by = hsda,
  from = "2025-12-29",
  to = "2026-01-26"
)
detection$expected <- round(detection$expected, 1)
detection[
  detection$hsda == "South Vancouver Island",
  c("period_start", "observed", "expected", "upper_limit", "alarm")
]
```

Install surveillance with `install.packages("surveillance", type = "binary")`
to use it. The rest of `islandepi` does not need it.

## Plot the counts

`islandepi` intentionally returns data rather than imposing a figure style. Use
`islandbrand::islh_epi_curve()` to create the Island Health figure:

```{r plot-setup, include=FALSE, eval=requireNamespace("islandbrand", quietly = TRUE)}
# R CMD check replays this code on the pdf device, which stops on a font
# family it does not know. islh_setup() picks a family that is installed.
islandbrand::islh_setup(format = "plots", quiet = TRUE)
```

```{r plot-example, eval=requireNamespace("islandbrand", quietly = TRUE), fig.width=8, fig.height=4.5, fig.alt="Stacked weekly case counts by health service delivery area from November 2025 to March 2026. Cases rise to a first peak in mid-December, dip over the holidays, peak again in late January and fall away by March. South Vancouver Island has the most cases each week."}
# Short labels keep the legend on one line.
weekly$area <- sub(" Vancouver Island", " Island", weekly$hsda)

islandbrand::islh_epi_curve(
  weekly,
  date = period_start,
  count = count,
  fill = area,
  labels = "total",
  title = "Weekly cases by health service delivery area",
  subtitle = "By week of symptom onset, November 2025 to March 2026",
  caption = "Source: simulated outbreak (islh_outbreak). Not real surveillance data."
) +
  ggplot2::labs(fill = NULL)
```

The published package website installs `islandbrand` and evaluates this example.
When this vignette is built on a computer without `islandbrand`, only the plot
is skipped; all analytical examples still run.

## Coverage and reporting calendar

Timestamp conversion defaults to `America/Vancouver`; supply `timezone` to
change the reporting calendar explicitly. Both the first and last incomplete
periods are excluded by default from counts, and a message says how many
events that left out. Preaggregated weekly counts cannot be relabelled to a
different week start.

Snapshots fail on missing source periods by default. Use
`missing_periods = "missing"` to propagate unknown values and `complete = FALSE`.
Use `missing_periods = "zero"` only after confirming those absent records mean
zero activity; that decision is recorded in `islh_screening`. Event-grid
completion likewise assumes that `from` and `to` describe confirmed coverage;
use `islh_check_coverage()` to confirm it.

Grouping columns must not share a name with a column the functions create,
such as `total`, `complete` or `upper_limit`. Rename them first; the functions
stop rather than overwrite the group labels.
